### 2023-06-20 manuscript/main.tex
added manuscript

%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[twocolumn]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{minted}
%% auto break lines
\setminted{breaklines=true}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2022}
\copyrightclause{
    Copyright for this paper by its authors.
    Use permitted under Creative Commons License Attribution 4.0 
    International (CC BY 4.0).
}

%%
%% This command is for the conference information
\conference{
    Woodstock'22: Symposium on the irreproducible science,
    June 07--11, 2022, Woodstock, NY
}

%%
%% The "title" command
\title{Prefix-free graphs and suffix array construction in sublinear space}

\tnotemark[1]
\tnotetext[1]{
    You can use this document as the template for preparing your
    publication. We recommend using the latest version of the ceurart style.
}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1,2]{Andrej Baláž}[%
orcid=0000-0002-0877-7063,
email=kulyabov-ds@rudn.ru,
url=https://yamadharma.github.io/,
]
\cormark[1]
\fnmark[1]
\address[1]{
    Peoples' Friendship University of Russia (RUDN University),
    6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation
}

%% Footnotes
\cortext[1]{Corresponding author.}
\fntext[1]{These authors contributed equally.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Recently, many graph structures for computational pangenomics were proposed.
  In order to be useful, these graph structures need to implement several
  operations such as efficient construction from many complete genomes,
  mapping of short and long read.
  These basic bioinformatics operations are well studied on sequential data,
  and together with data structures such as suffix trees, suffix arrays and 
  burrows wheeler transform allow high performance.
  Attempts to implement these operations on graphs brings with it many complications
  since these date structures are not easily obtainable for graphs.
  In this work, we introduce prefix-free graphs, a pangenomic data structure
  allowing to obtain the well known data structures from stringology in graph
  settings, which in turn allows for many efficient operations on pangenomes.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  LaTeX class \sep
  paper template \sep
  paper formatting \sep
  CEUR-WS
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Suffix array is a data structure used in many fields, such as stringology,
bioinformatics, and... .
Particularly in bioinformatics, it is used as part of FM index, which in turn
is used for read mapping.
Recently, a new paradigm in bioinformatics was formed, moving from a single
reference genome to a pangenome.
The term pangenome was first used by Tettelin et al. in 2005 during the
resequencing experiments on streptococcus.
Since then, the definition of a pangenome has shifted to any set of sequences
analysed jointly.
Nevertheless, most pangenomes consist of highly related sequences and thus are
highly repetitive.
This leads to representation of pangenomes as graphs, where similar genome
regions are represented by a single node and the nodes are conected to paths
representing the original sequences.
Several approaches of pangenomic graph construction exist, such as variation
graphs, cactus graphs, ...
Most of these approaches require an initial local alignment of similar regions,
which makes them computationally expensive.
Here we present a new class of graphs, prefix-free graphs, which are easy to
construct and provide a way how to efficiently index them to allow pattern 
search queries.
Pangenomes introduced by...
In computational pangenomics...
Repetitive datasets...
Graph pangenomes and textual pangenomes...
Prefix-free parsing is a technique...
Prefix-free graphs play an important role in improving space
complexity of several efficient algorithms, but are never the concept is never
framed as a pangenomic data structure.
We think that separating and clearly defining this data structure can lead to
more advancements...
Our contribution is in defining a data structure called prefix-free graphs and
showing how this data structure can be used to generate suffix array in
sublinear space.
Furthermore, we implemented two binarieš for an easy construction of prefix-free
graphs from a set of sequences, and from a graph in gfa format.
Furthermore, we implemented rust library for working with prefix-free graphs.
This library contains an iterator, which can be directly used to generate the 
suffix array in sublinear space.

\section{Prefix-free graphs}
The idea of prefix-free graphs is inspired by

\section{Suffix array construction}
\section{Related work}
MARIA \cite{2022maria}
Prefix-free Wheeler graphs \cite{2022pfwg}
Prefix-free parsing for building big BWTs \cite{2019boucher}
Rpair: rescaling repair with rsync \cite{2019gagie}

\section{Experiments}
\section{Discussion}

%%
%% The acknowledgments section is defined using the "acknowledgments" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acknowledgments}
  Thanks to the developers of ACM consolidated LaTeX styles
  \url{https://github.com/borisveytsman/acmart} and to the developers
  of Elsevier updated \LaTeX{} templates
  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.
\end{acknowledgments}

%%
%% Define the bibliography file to be used
\bibliography{sources}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Online Resources}

The sources for the ceur-art style are available via
\begin{itemize}
\item \href{https://github.com/yamadharma/ceurart}{GitHub},
% \item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
\item
  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/pkfscdkgkhcq}{Overleaf
    template}.
\end{itemize}

\end{document}

### 2023-06-20 manuscript/original.tex
added manuscript

%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[twocolumn]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{minted}
%% auto break lines
\setminted{breaklines=true}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2022}
\copyrightclause{
    Copyright for this paper by its authors.
    Use permitted under Creative Commons License Attribution 4.0 
    International (CC BY 4.0).
}

%%
%% This command is for the conference information
\conference{
    Woodstock'22: Symposium on the irreproducible science,
    June 07--11, 2022, Woodstock, NY
}

%%
%% The "title" command
\title{Prefix-free graphs and suffix array construction in sublinear space}

\tnotemark[1]
\tnotetext[1]{
    You can use this document as the template for preparing your
    publication. We recommend using the latest version of the ceurart style.
}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author[1,2]{Andrej Baláž}[%
orcid=0000-0002-0877-7063,
email=kulyabov-ds@rudn.ru,
url=https://yamadharma.github.io/,
]
\cormark[1]
\fnmark[1]
\address[1]{Peoples' Friendship University of Russia (RUDN University),
  6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation}
\address[2]{Joint Institute for Nuclear Research,
  6 Joliot-Curie, Dubna, Moscow region, 141980, Russian Federation}

\author[3]{Ilaria Tiddi}[%
orcid=0000-0001-7116-9338,
email=i.tiddi@vu.nl,
url=https://kmitd.github.io/ilaria/,
]
\fnmark[1]
\address[3]{Vrije Universiteit Amsterdam, De Boelelaan 1105, 1081 HV Amsterdam, The Netherlands}

\author[4]{Manfred Jeusfeld}[%
orcid=0000-0002-9421-8566,
email=Manfred.Jeusfeld@acm.org,
url=http://conceptbase.sourceforge.net/mjf/,
]
\fnmark[1]
\address[4]{University of Skövde, Högskolevägen 1, 541 28 Skövde, Sweden}

%% Footnotes
\cortext[1]{Corresponding author.}
\fntext[1]{These authors contributed equally.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX{} document is presented as an
  article formatted for publication by CEUR-WS in a conference
  proceedings. Based on the ``ceurart'' document class, this article
  presents and explains many of the common variations, as well as many
  of the formatting elements an author may use in the preparation of
  the documentation of their work.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  LaTeX class \sep
  paper template \sep
  paper formatting \sep
  CEUR-WS
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

CEUR-WS's article template provides a consistent \LaTeX{} style for
use across CEUR-WS publications, and incorporates accessibility and
metadata-extraction functionality. This document will explain the
major features of the document class.

If you are new to publishing with CEUR-WS, this document is a valuable
guide to the process of preparing your work for publication.

The ``\verb|ceurart|'' document class can be used to prepare articles
for any CEUR-WS publication, and for any stage of publication, from
review to final ``camera-ready'' copy with {\itshape very} few changes
to the source.

This class depends on the following packages
for its proper functioning:

\begin{itemize}
\item \verb|natbib.sty| for citation processing;
\item \verb|geometry.sty| for margin settings;
\item \verb|graphicx.sty| for graphics inclusion;
\item \verb|hyperref.sty| optional package if hyperlinking is required in
  the document;
\item \verb|fontawesome5.sty| optional package for bells and whistles.
\end{itemize}

All the above packages are part of any
standard \LaTeX{} installation.
Therefore, the users need not be
bothered about downloading any extra packages.

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \mintinline{latex}{\vspace} command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

\section{Template parameters}

There are a number of template
parameters which modify some part of the \verb|ceurart| document class.
This parameters are enclosed in square
brackets and are a part of the \mintinline{latex}{\documentclass} command:
\begin{minted}{latex}
  \documentclass[parameter]{ceurart}
\end{minted}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item \mintinline{latex}{twocolumn} : Two column layout.
\item \mintinline{latex}{hf} : Enable header and footer\footnote{You can enable
    the display of page numbers in the final version of the entire
    collection. In this case, you should adhere to the end-to-end
    pagination of individual papers.}.
\end{itemize}

\section{Front matter}

\subsection{Title Information}

The titles of papers should be either all use the emphasizing
capitalized style or they should all use the regular English (or
native language) style. It does not make a good impression if you or
your authors mix the styles.

Use the \mintinline{latex}{\title} command to define the title of your work. Do not
insert line breaks in your title.

\subsection{Title variants}

\mintinline{latex}{\title} command have the below options:
\begin{itemize}
\item \verb|title|: Document title. This is default option. 
\begin{minted}{latex}
\title[mode=title]{This is a title}
\end{minted}
You can just omit it, like as follows:
\begin{minted}{latex}
\title{This is a title}
\end{minted}

\item \verb|alt|: Alternate title.
\begin{minted}{latex}
\title[mode=alt]{This is a alternate title}
\end{minted}

\item \verb|sub|: Sub title.
\begin{minted}{latex}
\title[mode=sub]{This is a sub title}
\end{minted}
You can just use \mintinline{latex}{\subtitle} command, as follows:
\begin{minted}{latex}
\subtitle{This is a sub title}
\end{minted}

\item \verb|trans|: Translated title.
\begin{minted}{latex}
\title[mode=trans]{This is a translated title}
\end{minted}

\item \verb|transsub|: Translated sub title.
\begin{minted}{latex}
\title[mode=transsub]{This is a translated sub title}
\end{minted}
\end{itemize}

\subsection{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification. Multiple authors may share one affiliation. Authors'
names should not be abbreviated; use full first names wherever
possible. Include authors' e-mail addresses whenever possible.

\mintinline{latex}{\author} command have the below options:

\begin{itemize}
\item \mintinline{latex}{style} : Style of author name (chinese)
\item \mintinline{latex}{prefix} : Prefix
\item \mintinline{latex}{suffix} : Suffix
\item \mintinline{latex}{degree} : Degree
\item \mintinline{latex}{role} : Role
\item \mintinline{latex}{orcid} : ORCID
\item \mintinline{latex}{email} : E-mail
\item \mintinline{latex}{url} : URL
\end{itemize}

Author names can have some kinds of marks and notes:
\begin{itemize}
\item affiliation mark: \mintinline{latex}{\author[<num>]}.
\end{itemize}

The author names and affiliations could be formatted in two ways:
\begin{enumerate}
\item Group the authors per affiliation.
\item Use an explicit mark to indicate the affiliations.
\end{enumerate}

Author block example:
\begin{minted}{latex}
\author[1,2]{Author Name}[%
    prefix=Prof.,
    degree=D.Sc.,
    role=Researcher,
    orcid=0000-0000-000-0000,
    email=name@example.com,
    url=https://name.example.com
]

\address[1]{Affiliation #1}
\address[2]{Affiliation #2}
\end{minted}

\subsection{Abstract and Keywords}

Abstract shall be entered in an environment that starts
with \mintinline{latex}{\begin{abstract}} and ends with
\mintinline{latex}{\end{abstract}}.

\begin{minted}{latex}
\begin{abstract}
  This is an abstract.
\end{abstract}
\end{minted}

The key words are enclosed in a \mintinline{latex}{keywords}
environment. Use \mintinline{latex}{\sep} to separate keywords.

\begin{minted}{latex}
\begin{keywords}
  First keyword \sep 
  Second keyword \sep 
  Third keyword \sep 
  Fourth keyword
\end{keywords}
\end{minted}

At the end of front matter add \mintinline{latex}{\maketitle} command.

\subsection{Various Marks in the Front Matter}

The front matter becomes complicated due to various kinds
of notes and marks to the title and author names. Marks in
the title will be denoted by a star ($\star$) mark;
footnotes are denoted by super scripted Arabic numerals,
corresponding author by an Conformal asterisk (*) mark.

\subsubsection{Title marks}

Title mark can be entered by the command, \mintinline{latex}{\tnotemark[<num>]}
and the corresponding text can be entered with the command
\mintinline{latex}{\tnotetext[<num>]{<text>}}. An example will be:

\begin{minted}{latex}
\title{A better way to format your document for CEUR-WS}

\tnotemark[1]
\tnotetext[1]{You can use this document as the template for preparing your
  publication. We recommend using the latest version of the ceurart style.}
\end{minted}

\mintinline{latex}{\tnotemark} and \mintinline{latex}{\tnotetext} can be anywhere in
the front matter, but should be before \mintinline{latex}{\maketitle} command.

\subsubsection{Author marks}

Author names can have some kinds of marks and notes:
\begin{itemize}
\item footnote mark : \mintinline{latex}{\fnmark[<num>]}
\item footnote text : \mintinline{latex}{\fntext[<num>]{<text>}}
\item corresponding author mark : \mintinline{latex}{\cormark[<num>]}
\item corresponding author text : \mintinline{latex}{\cortext[<num>]{<text>}}
\end{itemize}

\subsubsection{Other marks}

At times, authors want footnotes which leave no marks in
the author names. The note text shall be listed as part of
the front matter notes. Class files provides
\mintinline{latex}{\nonumnote} for this purpose. The usage
\begin{minted}{latex}
\nonumnote{<text>}
\end{minted}
and should be entered anywhere before the \mintinline{latex}{\maketitle}
command for this to take effect.

\section{Sectioning Commands}

Your work should use standard \LaTeX{} sectioning commands:
\mintinline{latex}{\section}, \mintinline{latex}{\subsection},
\mintinline{latex}{\subsubsection}, and
\mintinline{latex}{\paragraph}. They should be numbered; do not remove
the numbering from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is not allowed.

\section{Tables}

The ``\verb|ceurart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed \textit{above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \verb|table| to enclose the table's contents and the
table caption. The contents of the table itself must go in the
\verb|tabular| environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table*}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table*}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \verb|table*| to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table}

\section{Math Equations}

You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}

A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \verb|math| environment,
which can be invoked with the usual
\mintinline{latex}{\begin} \ldots \mintinline{latex}{\end} construction or with
the short form \verb|$| \ldots \verb|$|. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX};
this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty} \frac{1}{n} = 0,
\end{math}
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}

A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \verb|equation|
environment. An unnumbered display equation is produced by the
\verb|displaymath| environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX{}; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty} \frac{1}{n} = 0.
\end{equation}
Notice how it is formatted somewhat differently in
the \verb|displaymath|
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  S_{n} = \sum_{i=1}^{n} x_{i} ,
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \lim_{x \to 0} (1 + x)^{1/x} = e
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
% \begin{figure}
%   \centering
%   \includegraphics[width=\linewidth]{}
%   \caption{1907 Franklin Model D roadster. Photograph by Harris \&
%     Ewing, Inc. [Public domain], via Wikimedia
%     Commons. (\url{https://goo.gl/VLCRBB}).}
% \end{figure}

Your figures should contain a caption which describes the figure to
the reader. Figure captions go below the figure. Your figures should
also include a description suitable for screen readers, to
assist the visually-challenged to better understand your work.

Figure captions are placed below the figure.

\section{Citations and Bibliographies}

The use of Bib\TeX{} for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \mintinline{latex}{\end{document}}
command:
\begin{minted}{latex}
\bibliography{bibfile}
\end{minted}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the Bib\TeX{} file.

\subsection{Some examples}

A paginated journal article \cite{Abril07}, an enumerated journal
article \cite{Cohen07}, a reference to an entire issue
\cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation
\cite{Editor00} followed by the same example, however we only output
the series if the volume number is given \cite{Editor00a} (so series
should not be present since it has no vol. no.), a chapter in a
divisible book \cite{Spector90}, a chapter in a divisible book in a
series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
an article in a proceedings (of a conference, symposium, workshop for
example) (paginated proceedings article) \cite{Andler79}, a
proceedings article with all possible elements \cite{Smith10}, an
example of an enumerated proceedings article \cite{VanGundy07}, an
informally published work \cite{Harel78}, a doctoral dissertation
\cite{Clarkson85}, a master's thesis: \cite{anisi03}, an online
document / world wide web resource \cite{Thornburg01, Ablamowicz07,
  Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2)
\cite{Novak03} and \cite{Lee05} and (Case 3) a patent
\cite{JoeScientist001}, work accepted for publication \cite{rous08},
prolific author \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other
cites might contain `duplicate' DOI and URLs (some SIAM articles)
\cite{Kirschmer:2010:AEI:1958016.1958018}. Multi-volume works as books
\cite{MR781536} and \cite{MR781537}. A couple of citations with DOIs:
\cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
citations: \cite{TUGInstmem, Thornburg01, R, UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{minted}{latex}
\begin{acknowledgments}
  These are different acknowledgments.
\end{acknowledgments}
\end{minted}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered
\mintinline{latex}{\section}; please use the ``\mintinline{latex}{acknowledgments}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\mintinline{latex}{\end{document}}'' command at the conclusion of your source
document.

Start the appendix with the ``\mintinline{latex}{\appendix}'' command:
\begin{minted}{latex}
\appendix
\end{minted}
and note that in the appendix, sections are lettered, not
numbered.

%%
%% The acknowledgments section is defined using the "acknowledgments" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acknowledgments}
  Thanks to the developers of ACM consolidated LaTeX styles
  \url{https://github.com/borisveytsman/acmart} and to the developers
  of Elsevier updated \LaTeX{} templates
  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.  
\end{acknowledgments}

%%
%% Define the bibliography file to be used
\bibliography{sample-ceur}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Online Resources}

The sources for the ceur-art style are available via
\begin{itemize}
\item \href{https://github.com/yamadharma/ceurart}{GitHub},
% \item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
\item
  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/pkfscdkgkhcq}{Overleaf
    template}.
\end{itemize}

\end{document}

%%
%% End of file

### 2023-06-20 src/bin/fasta2pfg.rs
adder source code

@@ -1,0 +1,66 @@
+use bio::io::fasta;
+use clap::Parser;
+use std::collections::HashMap;
+use std::fs::File;
+use maria::pf;
+
+/// Build prefix-free graph
+#[derive(Parser, Debug)]
+#[command(author, version, about)]
+struct Args {
+    /// fasta file
+    fasta_file: String,
+    /// trigger file, containing one trigger per line
+    trigger_file: String,
+    /// output in GFA format v1.1
+    output: String,
+}
+
+fn main() { 
+    let args = Args::parse();
+
+    let (trigs, trigs_size) = pf::load_trigs(&args.trigger_file);
+    let triggers = pf::get_triggers(&trigs, trigs_size);
+
+    let mut segments = HashMap::new();
+    let mut paths = Vec::new();
+
+    let fasta = File::open(args.fasta_file).expect("Cannot open fasta file.");
+    let mut records = fasta::Reader::new(fasta).records();
+    while let Some(Ok(record)) = records.next() {
+        // record.seq().len();
+        let mut seq = record.seq().to_owned();
+        seq.push(b'.');
+        pf::split_prefix_free(&seq, &triggers, &mut segments, &mut paths);
+    }
+
+    let (segments, paths) = pf::normalize(segments, paths);
+
+    let output = File::create(args.output).expect("Cannot create output file.");
+    pf::print_gfa(&segments, &paths, trigs_size, output)
+        .expect("Error writting GFA");
+}
+
+#[cfg(test)]
+mod tests {
+    use super::*;
+    use std::io::stdout;
+
+    #[test]
+    fn test_only_prefix_free() {
+        let triggers: Vec<&[u8]> = vec![b"T"];
+        let k = 1;
+        let seq1 = b"ATCTGTTAATG$";
+        let seq2 = b"AACGTGTACGTACGAAC$";
+
+        let mut segments = HashMap::new();
+        let mut paths = Vec::new();
+
+        pf::split_prefix_free(seq1, &triggers, &mut segments, &mut paths);
+        pf::split_prefix_free(seq2, &triggers, &mut segments, &mut paths);
+
+        let (segments, paths) = pf::normalize(segments, paths);
+
+        pf::print_gfa(&segments, &paths, k, stdout()).expect("Error writting GFA");
+    }
+}

### 2023-06-20 src/lib.rs

@@ -1,0 +1,22 @@
+pub mod arrays;
+pub mod naive;
+pub mod pf;
+
+fn reverse_complement(s: &[u8]) -> Vec<u8> {
+    let n = s.len();
+    let mut result = vec![0; n];
+    for (i, c) in s.iter().enumerate() {
+        match *c {
+            b'A' => { result[n-1-i] = b'T'; },
+            b'C' => { result[n-1-i] = b'G'; },
+            b'G' => { result[n-1-i] = b'C'; },
+            b'T' => { result[n-1-i] = b'A'; },
+            b'N' => { result[n-1-i] = b'N'; },
+            _ => { panic!("Unexpected letter in s."); }
+        }
+    }
+    return result;
+}
+
+#[cfg(test)]
+mod tests;

### 2023-06-20 src/pf.rs

@@ -1,0 +1,452 @@
+use bio::data_structures::suffix_array::lcp as lcp_array;
+use gfa::gfa::GFA;
+use std::io::Write;
+use std::io;
+use std::fs;
+use std::collections::HashMap;
+use gfa::gfa::Path;
+use gfa::gfa::Segment;
+use gfa::parser::GFAParser;
+use crate::arrays::SuffixArray;
+use std::ops::Add;
+use std::str;
+use crate::reverse_complement;
+
+/// Data required to iterate through suffix array in sublinear space
+pub struct PFData {
+    segment_join: Vec<u8>,
+    /// suffix array of segment join
+     sa: Vec<usize>, 
+    /// longest common prefix of segment join
+    lcp: Vec<isize>,
+    /// id of segments associated with suffix array
+     id: Vec<usize>,
+    /// position in the segment associated with suffix array
+    pos: Vec<usize>,
+
+    path_join: Vec<usize>,
+    /// length of segments
+    seg_len: Vec<usize>,
+    /// starting positions of a segment in expanded path join
+    seq_pos: Vec<Vec<usize>>,
+    /// rank of right contexts of path join
+    rc_rank: Vec<Vec<usize>>,
+
+    /// segment overlap
+    overlap: usize
+}
+
+pub struct PFDataIterator<'a> {
+    data: &'a PFData,
+    block: Block<'a>,
+}
+
+impl PFData {
+    pub fn new(segments: &[Vec<u8>], paths: &[Vec<usize>], overlap: usize) -> Self {
+        let segment_join = join(segments);
+
+        let sa  = SuffixArray::create(&*segment_join);
+        let lcp = lcp_array(&segment_join, &sa).decompress();
+        let isa = permutation_invert(&sa);
+        let id  = get_node_ids(&segment_join, &isa);
+        let pos = get_node_pos(&segment_join, &isa);
+
+        let path_join = join(paths);
+
+        let seg_len = get_lengths(segments);
+        let mut rc_rank = get_right_context_rank(&path_join, segments.len());
+        let mut seq_pos = get_sequence_position(&path_join, &seg_len, overlap);
+
+        for i in 0..segments.len() {
+            let iperm = argsort(&rc_rank[i]);
+            rc_rank[i] = permutation_apply(&iperm, &rc_rank[i]);
+            seq_pos[i] = permutation_apply(&iperm, &seq_pos[i]);
+        }
+
+        Self{
[…]

### 2023-06-20 src/bin/fasta2pfg.rs
fixed tests

@@ -2,7 +2,7 @@ use bio::io::fasta;
 use clap::Parser;
 use std::collections::HashMap;
 use std::fs::File;
-use maria::pf;
+use pfg::pf;
 
 /// Build prefix-free graph
 #[derive(Parser, Debug)]
@@ -28,7 +28,6 @@ fn main() {
     let fasta = File::open(args.fasta_file).expect("Cannot open fasta file.");
     let mut records = fasta::Reader::new(fasta).records();
     while let Some(Ok(record)) = records.next() {
-        // record.seq().len();
         let mut seq = record.seq().to_owned();
         seq.push(b'.');
         pf::split_prefix_free(&seq, &triggers, &mut segments, &mut paths);

### 2023-06-20 src/naive.rs

@@ -1,0 +1,103 @@
+use gfa::gfa::{Path, SegmentId, Segment};
+use gfa::optfields::*;
+use std::collections::HashMap;
+use std::hash::Hash;
+
+use crate::reverse_complement;
+
+fn parse_segments(s: &[u8]) -> Vec<(Vec<u8>, u8)> {
+    let mut start = 0;
+    let mut segments = Vec::new();
+    for i in 0..s.len() {
+        if s[i] == b',' {
+            let name = s[start..i-1].to_vec();
+            let sign = s[i-1];
+            segments.push((name, sign));
+            start = i+1;
+        }
+    }
+    let name = s[start..s.len()-1].to_vec();
+    let sign = s[s.len()-1];
+    segments.push((name, sign));
+    return segments;
+}
+
+fn get_sequence(
+    segments: &[(Vec<u8>, u8)], map: &HashMap<Vec<u8>, Vec<u8>>
+) -> Vec<u8> {
+    let mut sequence = Vec::new();
+    for (name, sign) in segments {
+        let node_string = map.get(name).unwrap();
+        if *sign == b'+' {
+            sequence.extend_from_slice(node_string);
+        } else {
+            let rc = reverse_complement(node_string);
+            sequence.extend_from_slice(&rc);
+        }
+    }
+    return sequence;
+}
+
+pub fn inverse_suffix_array(sa: &[usize]) -> Vec<usize> {
+    let mut isa = vec![0; sa.len()];
+    for i in 0..sa.len() { isa[sa[i]] = i; }
+    return isa;
+}
+
+pub fn tag_array<N: SegmentId, T: OptFields>(
+    sa: &[usize],
+    isa: &[usize],
+    paths: &[Path<N, T>],
+    map: &HashMap<Vec<u8>, Vec<u8>>
+) -> Vec<(Vec<u8>, usize)> {
+
+    let mut tag = vec![(Vec::new(), 0); sa.len()];
+    let mut j = 0;
+    for path in paths {
+        let path = parse_segments(&path.segment_names);
+        for (node_name, sign) in path {
+            let node = map.get(&node_name).unwrap();
+            for i in 0..node.len() {
+                let mut v = node_name.clone();
+                v.push(sign);
+                tag[isa[j]] = (v, i);
+                j += 1;
+            }
+        }
+        j += 1;
+    }
+    return tag;
+}
+
[…]

### 2023-06-20 src/bin/fasta2pfg.rs
fixed pfg generation

@@ -1,19 +1,16 @@
 use bio::io::fasta;
 use clap::Parser;
 use std::collections::HashMap;
-use std::fs::File;
 use pfg::pf;
+use std::io::{stdin, stdout};
 
 /// Build prefix-free graph
 #[derive(Parser, Debug)]
 #[command(author, version, about)]
 struct Args {
-    /// fasta file
-    fasta_file: String,
     /// trigger file, containing one trigger per line
+    #[arg(short)]
     trigger_file: String,
-    /// output in GFA format v1.1
-    output: String,
 }
 
 fn main() { 
@@ -25,18 +22,17 @@ fn main() {
     let mut segments = HashMap::new();
     let mut paths = Vec::new();
 
-    let fasta = File::open(args.fasta_file).expect("Cannot open fasta file.");
-    let mut records = fasta::Reader::new(fasta).records();
+    let mut records = fasta::Reader::new(stdin()).records();
     while let Some(Ok(record)) = records.next() {
         let mut seq = record.seq().to_owned();
-        seq.push(b'.');
+        let v = vec![b'.'; trigs_size];
+        seq.extend_from_slice(&v);
         pf::split_prefix_free(&seq, &triggers, &mut segments, &mut paths);
     }
 
     let (segments, paths) = pf::normalize(segments, paths);
 
-    let output = File::create(args.output).expect("Cannot create output file.");
-    pf::print_gfa(&segments, &paths, trigs_size, output)
+    pf::print_gfa(&segments, &paths, trigs_size, stdout())
         .expect("Error writting GFA");
 }
 

### 2023-06-20 src/tests.rs
added test generation

@@ -1,0 +1,24 @@
+use crate::pf::PFData;
+
+#[test]
+fn iter_works() {
+    let pfdata = PFData::from_pfgraph("example/pfg.gfa");
+
+    for (sa, id, pos) in pfdata.iter() {
+        println!("{sa}\t{id}\t{pos}");
+    }
+}
+
+#[test]
+fn show_pfdata() {
+    let pfdata = PFData::from_pfgraph("example/pfg.gfa");
+    pfdata.print();
+}
+
+#[test]
+fn can_load_arbitrary_graph() {
+    // let pfdata = PFData::from_graph(
+    //     "data/pftag/test_arbitrary.gfa",
+    //     "data/pftag/triggers.txt"
+    // );
+}

### 2023-06-20 manuscript/main.tex
added pfg construction

@@
-%% The first command in your LaTeX source must be the \documentclass command.
-%%
 %% Options:
@@
 
-%%
-%% end of the preamble, start of the body of the document source.
 \begin{document}
 
-%%
-%% Rights management information.
-%% CC-BY is default license.
-\copyrightyear{2022}
+\copyrightyear{2023}
 \copyrightclause{
@@
 
-%%
-%% This command is for the conference information
 \conference{
-    Woodstock'22: Symposium on the irreproducible science,
-    June 07--11, 2022, Woodstock, NY
+    ITAT'23: Information technologies -- Applications and Theory
+    September 22--26, 2023, Tatranské Matliare, Vysoké Tatry
 }
 
-%%
-%% The "title" command
 \title{Prefix-free graphs and suffix array construction in sublinear space}
 
-\tnotemark[1]
-\tnotetext[1]{
-    You can use this document as the template for preparing your
-    publication. We recommend using the latest version of the ceurart style.
+\author[1]{Andrej Baláž}[
+orcid=0000-0001-0002-0003,
+email=andrejbalaz001@gmail.com,
+]
+\cormark[1]
+\address[1]{
+    Faculty of Mathematics, Physics and Informatics,
+    Comenius University, Mlynská dolina F1, 842 48 Bratislava, Slovakia
 }
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-\author[1,2]{Andrej Baláž}[%
-orcid=0000-0002-0877-7063,
-email=kulyabov-ds@rudn.ru,
-url=https://yamadharma.github.io/,
-]
-\cormark[1]
-\fnmark[1]
-\address[1]{
-    Peoples' Friendship University of Russia (RUDN University),
-    6 Miklukho-Maklaya St, Moscow, 117198, Russian Federation
-}
-
-%% Footnotes
+% Footnotes
 \cortext[1]{Corresponding author.}
-\fntext[1]{These authors contributed equally.}
 
@@
 \begin{keywords}
-  LaTeX class \sep
-  paper template \sep
-  paper formatting \sep
-  CEUR-WS
+  pangenome \sep
+  suffix array
 \end{keywords}
@@
 \section{Prefix-free graphs}
-The idea of prefix-free graphs is inspired by
+% some intro?
+The idea of prefix-free graphs is inspired by rsync \cite{}.
+% what are prefix-free graphs
+Given a set of sequences representing a pangenome, we can partition each
+sequence into \emph{segment}s.
+These segments form nodes of the prefix-free graph.
+The edges of a graph represent the adjacencies of the segment pairs, which are 
+present in the original strings.
+The adjacent segments are overlapping by a fixed number of characters $k$.
+We will call these $k$-sized overlaps triggers.
+Another important feature of the segments is that they contain triggers only at
+the start or at the end of a segment.
+This feature guarantees that the set of segments is prefix-free, i.e. no
+segment is a prefix of another segment.
+% add sketch of a proof
+
+% how to construct prefix free graphs
+To illustrate how to partition the sequences into segments consider a set of
+strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
+Next, assume we will partition the sequences using a set of trigger words 
+$T = {AC, CG}$.
+
+In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
+In our case, we need to append $k$ sentinels to the sequences.
+The reasons for doing so will become clear in the suffix array construction section.
+
+Then, using a sliding window we can scan through the sequences and each time we
+encounter a trigger word, we create a new segment from the start of the previous
+trigger word to the end of the current trigger word.
+Two special cases happen at the beginning of a sequence, when no previous
+trigger was encountered, and at the end.
+In the former case, the segment starts simply at the start of a sequence and in
+the later case the segment ends at a sequence end.
+
+During the sequence scan, each unique segment is assigned an ID.
+The original sequence can then be represented as a list of IDs, also called 
+\emph{path}s.
+
+The result of this procedure is a set of segments and a list of paths.
+In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
+and the paths are $[  ]$.
+
+Because each segment is represented in a set $S$ only once, the sum of lengths
+of segments is usually much smaller than the length of original sequences.
+
+This set of segments, together with the list of paths are already a prefix-free
+graph, but in order to be useful for the next part, we need to normalize it.
+By normalizing, we mean to sort the segments lexicographically and relabel the
+paths accordingly.
+This will give us a list of segments and a list of paths which can be directly
+represented in the GFA format as shown in the Figure \ref{}.
+
+The reconstruction of the original sequence can then be done by expanding 
+the particular path, ignoring last $k$ characters of each segment.
 
 \section{Suffix array construction}
+
 \section{Related work}
@@
 
-%%
-%% The acknowledgments section is defined using the "acknowledgments" environment
-%% (and NOT an unnumbered section). This ensures the proper
-%% identification of the section in the article metadata, and the
-%% consistent spelling of the heading.
 \begin{acknowledgments}
-  Thanks to the developers of ACM consolidated LaTeX styles
-  \url{https://github.com/borisveytsman/acmart} and to the developers
-  of Elsevier updated \LaTeX{} templates
-  \url{https://www.ctan.org/tex-archive/macros/latex/contrib/els-cas-templates}.
+    Thanks to Alessia and Askar for their encouragement while developing prefix-free graphs.
 \end{acknowledgments}
 
-%%
-%% Define the bibliography file to be used
 \bibliography{sources}
 
-%%
-%% If your work has an appendix, this is the place to put it.
 \appendix
@@
 
-The sources for the ceur-art style are available via
+The sources for the prefix-free graphs are available via
 \begin{itemize}
-\item \href{https://github.com/yamadharma/ceurart}{GitHub},
-% \item \href{https://www.overleaf.com/project/5e76702c4acae70001d3bc87}{Overleaf},
-\item
-  \href{https://www.overleaf.com/latex/templates/template-for-submissions-to-ceur-workshop-proceedings-ceur-ws-dot-org/pkfscdkgkhcq}{Overleaf
-    template}.
+\item \href{https://github.com/andynet/pfg}{GitHub}.
 \end{itemize}

### 2023-06-20 manuscript/main.tex
added gfa figure

@@
 This will give us a list of segments and a list of paths which can be directly
-represented in the GFA format as shown in the Figure \ref{}.
+represented in the GFA format as shown in the Figure \ref{fig:gfa}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/pfg.png}
+    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
+    \label{fig:gfa}
+\end{figure}
 

### 2023-06-20 manuscript/main.tex
added suffix array intro

@@
 \section{Suffix array construction}
+Suffix array is a permutation of string positions which lexicographically
+sorts the suffixes of the string starting at that position.
+It is extremely powerful data structure with many applications in efficient
+algorithms on strings.
+Enhanced by additional data structures, such as for example longest common
+prefix (LCP) array, it can fully substitute suffix tree and solve problems,
+such as exact pattern matching, repeat finding, maximum exact matching (MEM)
+finding, document retrieval and many more.
+
+There exists several algorithms for suffix array construction in linear time
+\cite{} with several practical implementations \cite{}.
+Altough their time complexity is linear, these algorithms become bottlenecks in
+some application due to their linear space complexity.
+This is especially relevant in pangenomics, where the datasets often do not fit
+in the memory.
+Nonetheless, as we shown in previous sections, prefix-free graphs can represent
+the pangenome in substantialy smaller space and therefore can fit in computer
+memory.
+
+Here we show another crucial advantage of prefix-free graphs - their ability to
+output the suffix array of the original string iterativelly.
+This allows us to generate the suffix array one-by-one with the possibility
+to compress or use in subsequential computations directly, giving us the ability
+to use the rich theory of suffix arrays without ever storing them.
 

### 2023-06-20 manuscript/main.tex
added minted to results

@@
 %% Options:
+
 %% twocolumn : Two column layout.
@@
 
+\subsection{Iterator preparation}
+In order to prepare the prefix-free graph to generate the suffix array of
+original pangenome we need to create several data structures.
+First, we concatenate all the prefix-free segments into a single string using
+a separator $\#$ and append a sentinel $\$$.
+We will call this concatenation \emph{segment join}.
+
+Next, for the segment join we calculate the suffix array and the LCP array.
+For this we can use the linear time, linear space algorithms mentioned previously,
+since the segment join is assumed to be much smaller than the original pangenome.
+
+Next, we need to calculate the segment ID array.
+Next, for each suffix of a segment join, we need to calculate the segment ID
+and the segment position.
+The segment ID tells us that the current suffix of a segment join starts with a particular segment.
+
 \section{Related work}
@@
 
-\section{Experiments}
+\section{Results}
+We implemented the prefix-free graphs package in rust programming language.
+There are two binary crates and one library crate included in the package.
+Binary crates serve the purpose of creating prefix-free graphs from FASTA and from GFA formats.
+Library crate provides an interface for working with the prefix-free graphs,
+mainly an iterator of suffix array in sublinear space and linear time.
+
+The binary crates are fasta2pfg and gfa2pfg with usage:
+\begin{minted}{bash}
+fasta2pfg -t triggers.txt < pangenome.fna > pfg.gfa
+
+gfa2pfg -t triggers.txt < pangenome.gfa > pfg.gfa
+\end{minted}
+
+The library can be used from within rust programming language as follows:
+
+\begin{minted}{rust}
+let pfg = PFG::load("pfg.gfa");
+
+for (i, sa_i) in pfg.iter().enumerate() {
+    println!("{}\t{}", i, sa_i);
+}
+\end{minted}
+
 \section{Discussion}

### 2023-06-20 manuscript/main.tex
added some more figures
added some related work

@@
 segment is a prefix of another segment.
-% add sketch of a proof
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
+    \caption{Visual proof that the trigger words induce prefix-free segments.}
+    \label{fig:proof}
+\end{figure}
 
@@
     \centering
-    \includegraphics[width=\linewidth]{images/pfg.png}
+    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
     \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
@@
 \end{figure}
+
+% add graph figure
 
@@
 
+\begin{figure*}
+    \centering
+    \includegraphics[width=\linewidth]{images/segment_table.png}
+    \caption{}
+    \label{fig:segment_sa}
+\end{figure*}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/segment_metainfo.png}
+    \caption{}
+    \label{fig:segment_meta}
+\end{figure}
+
 \section{Related work}
+There are several proposed data structures implicitly relying on the structure
+of prefix-free graphs.
+
+Rpair: rescaling repair with rsync \cite{2019gagie}
+Prefix-free parsing for building big BWTs \cite{2019boucher}
+MONI and r-index \cite{}
+Prefix-free Wheeler graphs \cite{2022pfwg}
 MARIA \cite{2022maria}
-Prefix-free Wheeler graphs \cite{2022pfwg}
-Prefix-free parsing for building big BWTs \cite{2019boucher}
-Rpair: rescaling repair with rsync \cite{2019gagie}
+
+All of these works implement their own version of prefix-free graphs.
+We argue that separating prefix-free graphs as a standalone data structure can
+bring several benefits:
+
+\begin{itemize}
+    \item reduce the complexity of the presentation
+    \item allowing for better optimization of algorithms related to prefix-free graphs
+    \item supporting the theoretical research by clearly deliminating the relevant terms
+    \item allowing to bring prefix-free graphs closer to the biological data
+\end{itemize}
+
+Related to the last point, we remark that a possible future research can be in
+the proper choice of trigger words.
+As an example, we used stop codons in our experiments.
+This could create prefix-free graphs capturing some biologically relevant phenomena.
+Other options are recombination hotspots, highly repetitive elements, different
+binding sites and sites with increased breakage.
 

### 2023-06-21 manuscript/main.tex
added suffix table

@@
 
-Next, we need to calculate the segment ID array.
-Next, for each suffix of a segment join, we need to calculate the segment ID
-and the segment position.
-The segment ID tells us that the current suffix of a segment join starts with a particular segment.
+Next, for each suffix of a segment join, we need to calculate the corresponding
+segment ID and the segment position.
+These two arrays represent the current suffix of a segment join starts with a
+particular segment and the position in this segment.
+This can be computed using an inverse permutation of a suffix array of a segment
+join in linear time.
+To illustrate the procedure, consider the segment join of our running example in
+Figure \ref{fig:ids_and_positions}.
+Each position of the join can be assigned a segment ID and a position in current
+segment by linearly iterating the segment join and incrementing ID and position
+accordingly.
+Then, applying the inverse permutation of a suffix array to these arrays will
+change the order of computed values in a way corresponding to the sorted suffixes.
+The Table \ref{} shows the resulting suffix array, LCP array, segment ID array and segment position array.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
+    \caption{
+        Segment IDs and segment positions of a segment join. Permuting these
+        according to the inverse permutation of a suffix array results in 
+        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
+    }
+    \label{fig:ids_and_positions}
+\end{figure}
+
+\input{sa_table.tex}
 

### 2023-06-21 manuscript/main.tex
added segment table

@@
 change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{} shows the resulting suffix array, LCP array, segment ID array and segment position array.
+The Table \ref{tab:suffix} shows the resulting suffix array, LCP array, segment ID array and segment position array.
 
@@
 
-\begin{figure*}
-    \centering
-    \includegraphics[width=\linewidth]{images/segment_table.png}
-    \caption{}
-    \label{fig:segment_sa}
-\end{figure*}
+In the Table \ref{tab:suffix}, one row can represent multiple positions of the
+pangenome.
+In order to be able to reconstruct the suffix array, we need to store some
+additional information for each segment.
+For each segment, we store it's length, it's starting positions in the implicit
+concatenated pangenome and the rank of it's right context.
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/segment_metainfo.png}
-    \caption{}
-    \label{fig:segment_meta}
-\end{figure}
+\input{segment_table.tex}
 

### 2023-06-21 manuscript/main.tex
added path_join

@@
 We will call this concatenation \emph{segment join}.
+An example of a segment join is in the Figure \ref{fig:ids_and_positions}.
 
@@
 For this we can use the linear time, linear space algorithms mentioned previously,
-since the segment join is assumed to be much smaller than the original pangenome.
+since the segment join is usually much smaller than the original pangenome.
 
@@
 join in linear time.
-To illustrate the procedure, consider the segment join of our running example in
+To illustrate the procedure, consider the segment join of our running example from
 Figure \ref{fig:ids_and_positions}.
@@
 change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{tab:suffix} shows the resulting suffix array, LCP array, segment ID array and segment position array.
+The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
+segment ID array and segment position array.
 
@@
 pangenome.
-In order to be able to reconstruct the suffix array, we need to store some
-additional information for each segment.
-For each segment, we store it's length, it's starting positions in the implicit
-concatenated pangenome and the rank of it's right context.
+To identify these positions, we store some additional information in a second
+table \ref{tab:segment}.
+For each segment, we store it's length, it's starting positions in the pangenome
+and the rank of the right context of these positions.
+To calculate the starting positions and the ranks of the right contexts we use
+a path join.
+Similarly to segment join, a path join is a concatenation with delimiters $\#$
+and sentinel $\$$, but now constructed by concatenating the paths.
+An example of a path join for our running example is in the Figure \ref{fig:path_join}.
+Starting positions can then be calculated by cumulatively summing the lengths
+of the segments in path join and subtracting the overlaps.
+Computation of the ranks is more involved.
+We need to construct the suffix array of the path join and find it's inverse
+permutation $ISA$.
+This gives us ranks for each position in the path join.
+To determine the rank of the right context for position $i$, we simply take
+the value of the ranks $ISA[i+1]$.
+Finaly, we store the starts and ranks sorted by the rank values in the segment
+table as shown in Table \ref{tab:segment}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/path_join.png}
+    \caption{}
+    \label{fig:path_join}
+\end{figure}
 
 \input{segment_table.tex}
+
+\subsection{Iterating suffix array}
+With the previous tables in memory, we have all we need to generate the suffix
+array entries.
+Due to the prefix-freeness of the segments and the construction of the segment join
+we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
+Then, each row in the suffix table represents a suffix of a particular segment.
+
+There are five cases we need to consider.
+\begin{itemize}
+    \item suffix represents a separator or the sentinel
+    \item suffix represents a trigger or it's suffix
+    \item suffix longer than the trigger representing a single position in the pangenome
+    \item suffix longer than the trigger representing several positions in the pangenome
+\end{itemize}
+
+We can skip the first $n+1$ rows of the suffix table, because these suffixes
+do not yield any suffix array entry.
 

### 2023-06-21 manuscript/main.tex
added figures

@@
 reference genome to a pangenome.
-The term pangenome was first used by Tettelin et al. in 2005 during the
+The term pangenome was first used by \citet{tettelin2005genome} in 2005 during the
 resequencing experiments on streptococcus.
 Since then, the definition of a pangenome has shifted to any set of sequences
-analysed jointly.
+analysed jointly \cite{computational2018computational}.
 Nevertheless, most pangenomes consist of highly related sequences and thus are
@@
 Several approaches of pangenomic graph construction exist, such as variation
-graphs, cactus graphs, ...
+graphs \cite{church2015extending}?, cactus graphs \cite{paten2011cactus}, 
+Wheeler graphs \cite{gagie2017wheeler} ...
 Most of these approaches require an initial local alignment of similar regions,
@@
 sublinear space.
-Furthermore, we implemented two binarieš for an easy construction of prefix-free
+Furthermore, we implemented two binaries for an easy construction of prefix-free
 graphs from a set of sequences, and from a graph in gfa format.
@@
 % some intro?
-The idea of prefix-free graphs is inspired by rsync \cite{}.
+The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
+%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
+%% strings into blocks such that long repeated substrings are parsed the same
+%% way (except possibly at the beginning or end of the substrings).
+
 % what are prefix-free graphs
@@
     \centering
-    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
+    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
+    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
+    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
     \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}

### 2023-06-21 manuscript/main.tex
some more progress

@@
 \cormark[1]
+
+\author[1]{Alessia Petescia}[
+    orcid=0000,
+    email=alessia@gmail.com,
+]
+
 \address[1]{
+    Department of Applied Informatics,
     Faculty of Mathematics, Physics and Informatics,
-    Comenius University, Mlynská dolina F1, 842 48 Bratislava, Slovakia
+    Comenius University, Bratislava, Slovakia
 }
@@
 % Footnotes
-\cortext[1]{Corresponding author.}
+\cortext[1]{Corresponding author}
 
@@
 
-There are five cases we need to consider.
+There are three cases where a letter at position $SA[i]$ can occur in these suffix segments:
 \begin{itemize}
-    \item suffix represents a separator or the sentinel
-    \item suffix represents a trigger or it's suffix
-    \item suffix longer than the trigger representing a single position in the pangenome
-    \item suffix longer than the trigger representing several positions in the pangenome
+    \item it is a separator or the sentinel
+    \item it is at the end of a segment inside the trigger part
+    \item it is part of the segment outside the trigger part, separator and sentinel
 \end{itemize}
 
-We can skip the first $n+1$ rows of the suffix table, because these suffixes
-do not yield any suffix array entry.
+In the first case, there is no corresponding position in the original pangenome,
+therefore we can skip these rows of the suffix table.
+% We can skip the first $n+1$ rows of the suffix table, because these suffixes
+% do not yield any suffix array entry.
+
+In the second case, the position is inside the trigger string.
+Since the possitions inside the triggers are represented twice, once at the end
+of a segment a second time at the beginning of the following segment, we need
+to choose when to report them.
+A natural choice is to report positions at the beginning of a segment and ignore
+positions at the end.
+Here we note that this choice also plays nicely with the previous choice of sentinel
+characters, since ignoring last $k$ characters will result in not printing
+sentinels either.
+In conclusion, if the length of a current segment suffix is smaller or equal to
+the size of the trigger words, we skip the row similarly as in the previous case.
+
+Finaly, in the last case we report the positions.
+The last case will partition the suffix table into the blocks of the same suffixes.
+As an example consider the rows 20 and 21, which together for a single block.
+All other blocks in the example consist of single rows and therefore we call them singletons.
+
+This partitioning leads to three cases (from simplest to hardest):
+\begin{itemize}
+    \item A singleton block, occuring only once in the whole pangenome
+    \item A singleton block, repeating several times in the pangenome
+    \item A non-singleton block
+\end{itemize}
 
@@
 Prefix-free parsing for building big BWTs \cite{2019boucher}
-MONI and r-index \cite{}
+MONI and r-index \cite{rossi2022moni,gagie2020fully}
 Prefix-free Wheeler graphs \cite{2022pfwg}

### 2023-06-22 manuscript/main.tex
finished iteration

@@
 
+In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
+Given the row index $i$, this entry can be calculated with the following equation:
+
+\begin{equation}
+    \label{eq:sa_entry}
+    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
+\end{equation}
+
+As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
+It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
+The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
+This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.
+
+The second, slightly more complex case is a singleton block with several occurences in the pangenome.
+For this case, we need to report as many suffix array entries as the number of occurences.
+Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.
+
+As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
+This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
+Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.
+
+In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
+These suffixes represent the same substring of the original pangenome.
+Here, we report a suffix array entry for each occurence of the substring in the pangenome.
+To identify which entry is the first, we need to find the starting position with the smallest right context rank.
+Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
+To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.
+
+As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
+The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
+The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
+Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.
+
 \section{Related work}

### 2023-06-22 manuscript/comparison_table.tex
added experiments

\begin{table*}
    \label{tab:segment}
    \begin{tabular}{rllll}
        \toprule
        \#sequences & pggb & vg & minigraph-cactus & PFG \\
        \midrule
        % 2 &
        4   & 0m12s & 0m1s  & 3m47s  & 0m1s \\
        16  & 0m19s & 0m1s  & 6m30s  & 0m1s \\
        64  & 1m20s & 0m2s  & 13m30s & 0m1s \\
        256 & 7m20s & 0m55s & 41m49s & 0m3s \\
        \bottomrule
    \end{tabular}
    \caption{
        Comparison of running times of pangenomic graph construction tools.
        Some of these tools need some kind of preprocessing in order to build the graph.
        pggb requires for the fasta files to be indexed.
        In our case, due to the small size of our files, this time was negligible, but it can have a bigger impact when bigger genomes are considered.
        VG builds a pangenome graph starting from a VCF file or a multiple sequence alignment.
        In each case, those outputs require to perform time consuming steps.
        Having a suitable VCF file requires to align each haplotype to a reference genome and subsequently performing variant calling.
        Beyond being time consuming, using a linear reference in a first place can introduce some biases in the first place.
        On the other hand, performing a multiple sequence alignment is also known to be potentially slow, according to the length and the abundance of the underlying sequences.
    }
\end{table*}

### 2023-06-22 manuscript/introduction.tex
split manuscript into several parts

--- 2023-06-21 manuscript/main.tex
@@
-%% Options:
-
-%% twocolumn : Two column layout.
-%% hf: enable header and footer.
-\documentclass[twocolumn]{ceurart}
-
-%%
-%% One can fix some overfulls
-\sloppy
-
-%%
-%% Minted listings support 
-%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
-\usepackage{minted}
-%% auto break lines
-\setminted{breaklines=true}
-
-\begin{document}
-
-\copyrightyear{2023}
-\copyrightclause{
-    Copyright for this paper by its authors.
-    Use permitted under Creative Commons License Attribution 4.0 
-    International (CC BY 4.0).
-}
-
-\conference{
-    ITAT'23: Information technologies -- Applications and Theory
-    September 22--26, 2023, Tatranské Matliare, Vysoké Tatry
-}
-
-\title{Prefix-free graphs and suffix array construction in sublinear space}
-
-\author[1]{Andrej Baláž}[
-orcid=0000-0001-0002-0003,
-email=andrejbalaz001@gmail.com,
-]
-\cormark[1]
-\address[1]{
-    Faculty of Mathematics, Physics and Informatics,
-    Comenius University, Mlynská dolina F1, 842 48 Bratislava, Slovakia
-}
-
-% Footnotes
-\cortext[1]{Corresponding author.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
-\begin{abstract}
-  Recently, many graph structures for computational pangenomics were proposed.
-  In order to be useful, these graph structures need to implement several
-  operations such as efficient construction from many complete genomes,
-  mapping of short and long read.
-  These basic bioinformatics operations are well studied on sequential data,
-  and together with data structures such as suffix trees, suffix arrays and 
-  burrows wheeler transform allow high performance.
-  Attempts to implement these operations on graphs brings with it many complications
-  since these date structures are not easily obtainable for graphs.
-  In this work, we introduce prefix-free graphs, a pangenomic data structure
-  allowing to obtain the well known data structures from stringology in graph
-  settings, which in turn allows for many efficient operations on pangenomes.
-\end{abstract}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\begin{keywords}
-  pangenome \sep
-  suffix array
-\end{keywords}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
-
 \section{Introduction}
@@
 suffix array in sublinear space.
-
-\section{Prefix-free graphs}
-% some intro?
-The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
-%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
-%% strings into blocks such that long repeated substrings are parsed the same
-%% way (except possibly at the beginning or end of the substrings).
-
-% what are prefix-free graphs
-Given a set of sequences representing a pangenome, we can partition each
-sequence into \emph{segment}s.
-These segments form nodes of the prefix-free graph.
-The edges of a graph represent the adjacencies of the segment pairs, which are 
-present in the original strings.
-The adjacent segments are overlapping by a fixed number of characters $k$.
-We will call these $k$-sized overlaps triggers.
-Another important feature of the segments is that they contain triggers only at
-the start or at the end of a segment.
-This feature guarantees that the set of segments is prefix-free, i.e. no
-segment is a prefix of another segment.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
-    \caption{Visual proof that the trigger words induce prefix-free segments.}
-    \label{fig:proof}
-\end{figure}
-
-% how to construct prefix free graphs
-To illustrate how to partition the sequences into segments consider a set of
-strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
-Next, assume we will partition the sequences using a set of trigger words 
-$T = {AC, CG}$.
-
-In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
-In our case, we need to append $k$ sentinels to the sequences.
-The reasons for doing so will become clear in the suffix array construction section.
-
-Then, using a sliding window we can scan through the sequences and each time we
-encounter a trigger word, we create a new segment from the start of the previous
-trigger word to the end of the current trigger word.
-Two special cases happen at the beginning of a sequence, when no previous
-trigger was encountered, and at the end.
-In the former case, the segment starts simply at the start of a sequence and in
-the later case the segment ends at a sequence end.
-
-During the sequence scan, each unique segment is assigned an ID.
-The original sequence can then be represented as a list of IDs, also called 
-\emph{path}s.
-
-The result of this procedure is a set of segments and a list of paths.
-In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
-and the paths are $[  ]$.
-
-Because each segment is represented in a set $S$ only once, the sum of lengths
-of segments is usually much smaller than the length of original sequences.
-
-This set of segments, together with the list of paths are already a prefix-free
-graph, but in order to be useful for the next part, we need to normalize it.
-By normalizing, we mean to sort the segments lexicographically and relabel the
-paths accordingly.
-This will give us a list of segments and a list of paths which can be directly
-represented in the GFA format as shown in the Figure \ref{fig:gfa}.
-
-\begin{figure}
-    \centering
-    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
-    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
-    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
-    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
-    \label{fig:gfa}
-\end{figure}
-
-% add graph figure
-
-The reconstruction of the original sequence can then be done by expanding 
-the particular path, ignoring last $k$ characters of each segment.
-
-\section{Suffix array construction}
-Suffix array is a permutation of string positions which lexicographically
-sorts the suffixes of the string starting at that position.
-It is extremely powerful data structure with many applications in efficient
-algorithms on strings.
-Enhanced by additional data structures, such as for example longest common
-prefix (LCP) array, it can fully substitute suffix tree and solve problems,
-such as exact pattern matching, repeat finding, maximum exact matching (MEM)
-finding, document retrieval and many more.
-
-There exists several algorithms for suffix array construction in linear time
-\cite{} with several practical implementations \cite{}.
-Altough their time complexity is linear, these algorithms become bottlenecks in
-some application due to their linear space complexity.
-This is especially relevant in pangenomics, where the datasets often do not fit
-in the memory.
-Nonetheless, as we shown in previous sections, prefix-free graphs can represent
-the pangenome in substantialy smaller space and therefore can fit in computer
-memory.
-
-Here we show another crucial advantage of prefix-free graphs - their ability to
-output the suffix array of the original string iterativelly.
-This allows us to generate the suffix array one-by-one with the possibility
-to compress or use in subsequential computations directly, giving us the ability
-to use the rich theory of suffix arrays without ever storing them.
-
-\subsection{Iterator preparation}
-In order to prepare the prefix-free graph to generate the suffix array of
-original pangenome we need to create several data structures.
-First, we concatenate all the prefix-free segments into a single string using
-a separator $\#$ and append a sentinel $\$$.
-We will call this concatenation \emph{segment join}.
-An example of a segment join is in the Figure \ref{fig:ids_and_positions}.
-
-Next, for the segment join we calculate the suffix array and the LCP array.
-For this we can use the linear time, linear space algorithms mentioned previously,
-since the segment join is usually much smaller than the original pangenome.
-
-Next, for each suffix of a segment join, we need to calculate the corresponding
-segment ID and the segment position.
-These two arrays represent the current suffix of a segment join starts with a
-particular segment and the position in this segment.
-This can be computed using an inverse permutation of a suffix array of a segment
-join in linear time.
-To illustrate the procedure, consider the segment join of our running example from
-Figure \ref{fig:ids_and_positions}.
-Each position of the join can be assigned a segment ID and a position in current
-segment by linearly iterating the segment join and incrementing ID and position
-accordingly.
-Then, applying the inverse permutation of a suffix array to these arrays will
-change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
-segment ID array and segment position array.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
-    \caption{
-        Segment IDs and segment positions of a segment join. Permuting these
-        according to the inverse permutation of a suffix array results in 
-        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
-    }
-    \label{fig:ids_and_positions}
-\end{figure}
-
-\input{sa_table.tex}
-
-In the Table \ref{tab:suffix}, one row can represent multiple positions of the
-pangenome.
-To identify these positions, we store some additional information in a second
-table \ref{tab:segment}.
-For each segment, we store it's length, it's starting positions in the pangenome
-and the rank of the right context of these positions.
-To calculate the starting positions and the ranks of the right contexts we use
-a path join.
-Similarly to segment join, a path join is a concatenation with delimiters $\#$
-and sentinel $\$$, but now constructed by concatenating the paths.
-An example of a path join for our running example is in the Figure \ref{fig:path_join}.
-Starting positions can then be calculated by cumulatively summing the lengths
-of the segments in path join and subtracting the overlaps.
-Computation of the ranks is more involved.
-We need to construct the suffix array of the path join and find it's inverse
-permutation $ISA$.
-This gives us ranks for each position in the path join.
-To determine the rank of the right context for position $i$, we simply take
-the value of the ranks $ISA[i+1]$.
-Finaly, we store the starts and ranks sorted by the rank values in the segment
-table as shown in Table \ref{tab:segment}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/path_join.png}
-    \caption{}
-    \label{fig:path_join}
-\end{figure}
-
-\input{segment_table.tex}
-
-\subsection{Iterating suffix array}
-With the previous tables in memory, we have all we need to generate the suffix
-array entries.
-Due to the prefix-freeness of the segments and the construction of the segment join
-we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
-Then, each row in the suffix table represents a suffix of a particular segment.
-
-There are five cases we need to consider.
-\begin{itemize}
-    \item suffix represents a separator or the sentinel
-    \item suffix represents a trigger or it's suffix
-    \item suffix longer than the trigger representing a single position in the pangenome
-    \item suffix longer than the trigger representing several positions in the pangenome
-\end{itemize}
-
-We can skip the first $n+1$ rows of the suffix table, because these suffixes
-do not yield any suffix array entry.
-
-\section{Related work}
-There are several proposed data structures implicitly relying on the structure
-of prefix-free graphs.
-
-Rpair: rescaling repair with rsync \cite{2019gagie}
-Prefix-free parsing for building big BWTs \cite{2019boucher}
-MONI and r-index \cite{}
-Prefix-free Wheeler graphs \cite{2022pfwg}
-MARIA \cite{2022maria}
-
-All of these works implement their own version of prefix-free graphs.
-We argue that separating prefix-free graphs as a standalone data structure can
-bring several benefits:
-
-\begin{itemize}
-    \item reduce the complexity of the presentation
-    \item allowing for better optimization of algorithms related to prefix-free graphs
-    \item supporting the theoretical research by clearly deliminating the relevant terms
-    \item allowing to bring prefix-free graphs closer to the biological data
-\end{itemize}
-
-Related to the last point, we remark that a possible future research can be in
-the proper choice of trigger words.
-As an example, we used stop codons in our experiments.
-This could create prefix-free graphs capturing some biologically relevant phenomena.
-Other options are recombination hotspots, highly repetitive elements, different
-binding sites and sites with increased breakage.
-
-\section{Results}
-We implemented the prefix-free graphs package in rust programming language.
-There are two binary crates and one library crate included in the package.
-Binary crates serve the purpose of creating prefix-free graphs from FASTA and from GFA formats.
-Library crate provides an interface for working with the prefix-free graphs,
-mainly an iterator of suffix array in sublinear space and linear time.
-
-The binary crates are fasta2pfg and gfa2pfg with usage:
-\begin{minted}{bash}
-fasta2pfg -t triggers.txt < pangenome.fna > pfg.gfa
-
-gfa2pfg -t triggers.txt < pangenome.gfa > pfg.gfa
-\end{minted}
-
-The library can be used from within rust programming language as follows:
-
-\begin{minted}{rust}
-let pfg = PFG::load("pfg.gfa");
-
-for (i, sa_i) in pfg.iter().enumerate() {
-    println!("{}\t{}", i, sa_i);
-}
-\end{minted}
-
-\section{Discussion}
-
-\begin{acknowledgments}
-    Thanks to Alessia and Askar for their encouragement while developing prefix-free graphs.
-\end{acknowledgments}
-
-\bibliography{sources}
-
-\appendix
-
-\section{Online Resources}
-
-The sources for the prefix-free graphs are available via
-\begin{itemize}
-\item \href{https://github.com/andynet/pfg}{GitHub}.
-\end{itemize}
-
-\end{document}

### 2023-06-22 manuscript/main.tex
added experiments
split manuscript into several parts

--- 2023-06-21 manuscript/main.tex
@@
 
-\section{Introduction}
-Suffix array is a data structure used in many fields, such as stringology,
-bioinformatics, and... .
-Particularly in bioinformatics, it is used as part of FM index, which in turn
-is used for read mapping.
-Recently, a new paradigm in bioinformatics was formed, moving from a single
-reference genome to a pangenome.
-The term pangenome was first used by \citet{tettelin2005genome} in 2005 during the
-resequencing experiments on streptococcus.
-Since then, the definition of a pangenome has shifted to any set of sequences
-analysed jointly \cite{computational2018computational}.
-Nevertheless, most pangenomes consist of highly related sequences and thus are
-highly repetitive.
-This leads to representation of pangenomes as graphs, where similar genome
-regions are represented by a single node and the nodes are conected to paths
-representing the original sequences.
-Several approaches of pangenomic graph construction exist, such as variation
-graphs \cite{church2015extending}?, cactus graphs \cite{paten2011cactus}, 
-Wheeler graphs \cite{gagie2017wheeler} ...
-Most of these approaches require an initial local alignment of similar regions,
-which makes them computationally expensive.
-Here we present a new class of graphs, prefix-free graphs, which are easy to
-construct and provide a way how to efficiently index them to allow pattern 
-search queries.
-Pangenomes introduced by...
-In computational pangenomics...
-Repetitive datasets...
-Graph pangenomes and textual pangenomes...
-Prefix-free parsing is a technique...
-Prefix-free graphs play an important role in improving space
-complexity of several efficient algorithms, but are never the concept is never
-framed as a pangenomic data structure.
-We think that separating and clearly defining this data structure can lead to
-more advancements...
-Our contribution is in defining a data structure called prefix-free graphs and
-showing how this data structure can be used to generate suffix array in
-sublinear space.
-Furthermore, we implemented two binaries for an easy construction of prefix-free
-graphs from a set of sequences, and from a graph in gfa format.
-Furthermore, we implemented rust library for working with prefix-free graphs.
-This library contains an iterator, which can be directly used to generate the 
-suffix array in sublinear space.
+\input{introduction.tex}
 
-\section{Prefix-free graphs}
-% some intro?
-The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
-%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
-%% strings into blocks such that long repeated substrings are parsed the same
-%% way (except possibly at the beginning or end of the substrings).
+\input{pfg.tex}
 
-% what are prefix-free graphs
-Given a set of sequences representing a pangenome, we can partition each
-sequence into \emph{segment}s.
-These segments form nodes of the prefix-free graph.
-The edges of a graph represent the adjacencies of the segment pairs, which are 
-present in the original strings.
-The adjacent segments are overlapping by a fixed number of characters $k$.
-We will call these $k$-sized overlaps triggers.
-Another important feature of the segments is that they contain triggers only at
-the start or at the end of a segment.
-This feature guarantees that the set of segments is prefix-free, i.e. no
-segment is a prefix of another segment.
+\input{sa_construction.tex}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
-    \caption{Visual proof that the trigger words induce prefix-free segments.}
-    \label{fig:proof}
-\end{figure}
+\input{results.tex}
 
-% how to construct prefix free graphs
-To illustrate how to partition the sequences into segments consider a set of
-strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
-Next, assume we will partition the sequences using a set of trigger words 
-$T = {AC, CG}$.
-
-In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
-In our case, we need to append $k$ sentinels to the sequences.
-The reasons for doing so will become clear in the suffix array construction section.
-
-Then, using a sliding window we can scan through the sequences and each time we
-encounter a trigger word, we create a new segment from the start of the previous
-trigger word to the end of the current trigger word.
-Two special cases happen at the beginning of a sequence, when no previous
-trigger was encountered, and at the end.
-In the former case, the segment starts simply at the start of a sequence and in
-the later case the segment ends at a sequence end.
-
-During the sequence scan, each unique segment is assigned an ID.
-The original sequence can then be represented as a list of IDs, also called 
-\emph{path}s.
-
-The result of this procedure is a set of segments and a list of paths.
-In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
-and the paths are $[  ]$.
-
-Because each segment is represented in a set $S$ only once, the sum of lengths
-of segments is usually much smaller than the length of original sequences.
-
-This set of segments, together with the list of paths are already a prefix-free
-graph, but in order to be useful for the next part, we need to normalize it.
-By normalizing, we mean to sort the segments lexicographically and relabel the
-paths accordingly.
-This will give us a list of segments and a list of paths which can be directly
-represented in the GFA format as shown in the Figure \ref{fig:gfa}.
-
-\begin{figure}
-    \centering
-    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
-    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
-    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
-    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
-    \label{fig:gfa}
-\end{figure}
-
-% add graph figure
-
-The reconstruction of the original sequence can then be done by expanding 
-the particular path, ignoring last $k$ characters of each segment.
-
-\section{Suffix array construction}
-Suffix array is a permutation of string positions which lexicographically
-sorts the suffixes of the string starting at that position.
-It is extremely powerful data structure with many applications in efficient
-algorithms on strings.
-Enhanced by additional data structures, such as for example longest common
-prefix (LCP) array, it can fully substitute suffix tree and solve problems,
-such as exact pattern matching, repeat finding, maximum exact matching (MEM)
-finding, document retrieval and many more.
-
-There exists several algorithms for suffix array construction in linear time
-\cite{} with several practical implementations \cite{}.
-Altough their time complexity is linear, these algorithms become bottlenecks in
-some application due to their linear space complexity.
-This is especially relevant in pangenomics, where the datasets often do not fit
-in the memory.
-Nonetheless, as we shown in previous sections, prefix-free graphs can represent
-the pangenome in substantialy smaller space and therefore can fit in computer
-memory.
-
-Here we show another crucial advantage of prefix-free graphs - their ability to
-output the suffix array of the original string iterativelly.
-This allows us to generate the suffix array one-by-one with the possibility
-to compress or use in subsequential computations directly, giving us the ability
-to use the rich theory of suffix arrays without ever storing them.
-
-\subsection{Iterator preparation}
-In order to prepare the prefix-free graph to generate the suffix array of
-original pangenome we need to create several data structures.
-First, we concatenate all the prefix-free segments into a single string using
-a separator $\#$ and append a sentinel $\$$.
-We will call this concatenation \emph{segment join}.
-An example of a segment join is in the Figure \ref{fig:ids_and_positions}.
-
-Next, for the segment join we calculate the suffix array and the LCP array.
-For this we can use the linear time, linear space algorithms mentioned previously,
-since the segment join is usually much smaller than the original pangenome.
-
-Next, for each suffix of a segment join, we need to calculate the corresponding
-segment ID and the segment position.
-These two arrays represent the current suffix of a segment join starts with a
-particular segment and the position in this segment.
-This can be computed using an inverse permutation of a suffix array of a segment
-join in linear time.
-To illustrate the procedure, consider the segment join of our running example from
-Figure \ref{fig:ids_and_positions}.
-Each position of the join can be assigned a segment ID and a position in current
-segment by linearly iterating the segment join and incrementing ID and position
-accordingly.
-Then, applying the inverse permutation of a suffix array to these arrays will
-change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
-segment ID array and segment position array.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
-    \caption{
-        Segment IDs and segment positions of a segment join. Permuting these
-        according to the inverse permutation of a suffix array results in 
-        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
-    }
-    \label{fig:ids_and_positions}
-\end{figure}
-
-\input{sa_table.tex}
-
-In the Table \ref{tab:suffix}, one row can represent multiple positions of the
-pangenome.
-To identify these positions, we store some additional information in a second
-table \ref{tab:segment}.
-For each segment, we store it's length, it's starting positions in the pangenome
-and the rank of the right context of these positions.
-To calculate the starting positions and the ranks of the right contexts we use
-a path join.
-Similarly to segment join, a path join is a concatenation with delimiters $\#$
-and sentinel $\$$, but now constructed by concatenating the paths.
-An example of a path join for our running example is in the Figure \ref{fig:path_join}.
-Starting positions can then be calculated by cumulatively summing the lengths
-of the segments in path join and subtracting the overlaps.
-Computation of the ranks is more involved.
-We need to construct the suffix array of the path join and find it's inverse
-permutation $ISA$.
-This gives us ranks for each position in the path join.
-To determine the rank of the right context for position $i$, we simply take
-the value of the ranks $ISA[i+1]$.
-Finaly, we store the starts and ranks sorted by the rank values in the segment
-table as shown in Table \ref{tab:segment}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/path_join.png}
-    \caption{}
-    \label{fig:path_join}
-\end{figure}
-
-\input{segment_table.tex}
-
-\subsection{Iterating suffix array}
-With the previous tables in memory, we have all we need to generate the suffix
-array entries.
-Due to the prefix-freeness of the segments and the construction of the segment join
-we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
-Then, each row in the suffix table represents a suffix of a particular segment.
-
-There are five cases we need to consider.
-\begin{itemize}
-    \item suffix represents a separator or the sentinel
-    \item suffix represents a trigger or it's suffix
-    \item suffix longer than the trigger representing a single position in the pangenome
-    \item suffix longer than the trigger representing several positions in the pangenome
-\end{itemize}
-
-We can skip the first $n+1$ rows of the suffix table, because these suffixes
-do not yield any suffix array entry.
-
-\section{Related work}
-There are several proposed data structures implicitly relying on the structure
-of prefix-free graphs.
-
-Rpair: rescaling repair with rsync \cite{2019gagie}
-Prefix-free parsing for building big BWTs \cite{2019boucher}
-MONI and r-index \cite{}
-Prefix-free Wheeler graphs \cite{2022pfwg}
-MARIA \cite{2022maria}
-
-All of these works implement their own version of prefix-free graphs.
-We argue that separating prefix-free graphs as a standalone data structure can
-bring several benefits:
-
-\begin{itemize}
-    \item reduce the complexity of the presentation
-    \item allowing for better optimization of algorithms related to prefix-free graphs
-    \item supporting the theoretical research by clearly deliminating the relevant terms
-    \item allowing to bring prefix-free graphs closer to the biological data
-\end{itemize}
-
-Related to the last point, we remark that a possible future research can be in
-the proper choice of trigger words.
-As an example, we used stop codons in our experiments.
-This could create prefix-free graphs capturing some biologically relevant phenomena.
-Other options are recombination hotspots, highly repetitive elements, different
-binding sites and sites with increased breakage.
-
-\section{Results}
-We implemented the prefix-free graphs package in rust programming language.
-There are two binary crates and one library crate included in the package.
-Binary crates serve the purpose of creating prefix-free graphs from FASTA and from GFA formats.
-Library crate provides an interface for working with the prefix-free graphs,
-mainly an iterator of suffix array in sublinear space and linear time.
-
-The binary crates are fasta2pfg and gfa2pfg with usage:
-\begin{minted}{bash}
-fasta2pfg -t triggers.txt < pangenome.fna > pfg.gfa
-
-gfa2pfg -t triggers.txt < pangenome.gfa > pfg.gfa
-\end{minted}
-
-The library can be used from within rust programming language as follows:
-
-\begin{minted}{rust}
-let pfg = PFG::load("pfg.gfa");
-
-for (i, sa_i) in pfg.iter().enumerate() {
-    println!("{}\t{}", i, sa_i);
-}
-\end{minted}
-
-\section{Discussion}
+\input{conclusion.tex}
 
 \begin{acknowledgments}
-    Thanks to Alessia and Askar for their encouragement while developing prefix-free graphs.
+    Thanks to Askar for his encouragement while developing prefix-free graphs.
+    Thanks to Travis for introducing us to the advanced stringology.
+    Thanks to Adrian for discussing some ideas from this paper with us.
 \end{acknowledgments}

### 2023-06-22 manuscript/pfg.tex
split manuscript into several parts

\section{Prefix-free graphs}
% some intro?
The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
%% strings into blocks such that long repeated substrings are parsed the same
%% way (except possibly at the beginning or end of the substrings).

% what are prefix-free graphs
Given a set of sequences representing a pangenome, we can partition each
sequence into \emph{segment}s.
These segments form nodes of the prefix-free graph.
The edges of a graph represent the adjacencies of the segment pairs, which are 
present in the original strings.
The adjacent segments are overlapping by a fixed number of characters $k$.
We will call these $k$-sized overlaps triggers.
Another important feature of the segments is that they contain triggers only at
the start or at the end of a segment.
This feature guarantees that the set of segments is prefix-free, i.e. no
segment is a prefix of another segment.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
    \caption{Visual proof that the trigger words induce prefix-free segments.}
    \label{fig:proof}
\end{figure}

% how to construct prefix free graphs
To illustrate how to partition the sequences into segments consider a set of
strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
Next, assume we will partition the sequences using a set of trigger words 
$T = {AC, CG}$.

In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
In our case, we need to append $k$ sentinels to the sequences.
The reasons for doing so will become clear in the suffix array construction section.

Then, using a sliding window we can scan through the sequences and each time we
encounter a trigger word, we create a new segment from the start of the previous
trigger word to the end of the current trigger word.
Two special cases happen at the beginning of a sequence, when no previous
trigger was encountered, and at the end.
In the former case, the segment starts simply at the start of a sequence and in
the later case the segment ends at a sequence end.

During the sequence scan, each unique segment is assigned an ID.
The original sequence can then be represented as a list of IDs, also called 
\emph{path}s.

The result of this procedure is a set of segments and a list of paths.
In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
and the paths are $[  ]$.

Because each segment is represented in a set $S$ only once, the sum of lengths
of segments is usually much smaller than the length of original sequences.

This set of segments, together with the list of paths are already a prefix-free
graph, but in order to be useful for the next part, we need to normalize it.
By normalizing, we mean to sort the segments lexicographically and relabel the
paths accordingly.
This will give us a list of segments and a list of paths which can be directly
represented in the GFA format as shown in the Figure \ref{fig:gfa}.

\begin{figure}
    \centering
    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
    \label{fig:gfa}
\end{figure}

% add graph figure

The reconstruction of the original sequence can then be done by expanding 
the particular path, ignoring last $k$ characters of each segment.

### 2023-06-22 manuscript/results.tex
split manuscript into several parts

\section{Results}
We implemented the prefix-free graphs package in rust programming language.
There are two binary crates and one library crate included in the package.
Binary crates serve the purpose of creating prefix-free graphs from FASTA and from GFA formats.
Library crate provides an interface for working with the prefix-free graphs,
mainly an iterator of suffix array in sublinear space and linear time.

The binary crates are fasta2pfg and gfa2pfg with usage:
\begin{minted}{bash}
fasta2pfg -t triggers.txt < pangenome.fna > pfg.gfa

gfa2pfg -t triggers.txt < pangenome.gfa > pfg.gfa
\end{minted}

The library can be used from within rust programming language as follows:

\begin{minted}{rust}
let pfg = PFG::load("pfg.gfa");

for (i, sa_i) in pfg.iter().enumerate() {
    println!("{}\t{}", i, sa_i);
}
\end{minted}

We compared the pangenome construction of several tools.

\input{comparison_table.tex}

### 2023-06-22 manuscript/sa_construction.tex
split manuscript into several parts

\section{Suffix array construction}
Suffix array is a permutation of string positions which lexicographically
sorts the suffixes of the string starting at that position.
It is extremely powerful data structure with many applications in efficient
algorithms on strings.
Enhanced by additional data structures, such as for example longest common
prefix (LCP) array, it can fully substitute suffix tree and solve problems,
such as exact pattern matching, repeat finding, maximum exact matching (MEM)
finding, document retrieval and many more.

There exists several algorithms for suffix array construction in linear time
\cite{} with several practical implementations \cite{}.
Altough their time complexity is linear, these algorithms become bottlenecks in
some application due to their linear space complexity.
This is especially relevant in pangenomics, where the datasets often do not fit
in the memory.
Nonetheless, as we shown in previous sections, prefix-free graphs can represent
the pangenome in substantialy smaller space and therefore can fit in computer
memory.

Here we show another crucial advantage of prefix-free graphs - their ability to
output the suffix array of the original string iterativelly.
This allows us to generate the suffix array one-by-one with the possibility
to compress or use in subsequential computations directly, giving us the ability
to use the rich theory of suffix arrays without ever storing them.

\subsection{Iterator preparation}
In order to prepare the prefix-free graph to generate the suffix array of
original pangenome we need to create several data structures.
First, we concatenate all the prefix-free segments into a single string using
a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in the Figure \ref{fig:ids_and_positions}.

Next, for the segment join we calculate the suffix array and the LCP array.
For this we can use the linear time, linear space algorithms mentioned previously,
since the segment join is usually much smaller than the original pangenome.

Next, for each suffix of a segment join, we need to calculate the corresponding
segment ID and the segment position.
These two arrays represent the current suffix of a segment join starts with a
particular segment and the position in this segment.
This can be computed using an inverse permutation of a suffix array of a segment
join in linear time.
To illustrate the procedure, consider the segment join of our running example from
Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in current
segment by linearly iterating the segment join and incrementing ID and position
accordingly.
Then, applying the inverse permutation of a suffix array to these arrays will
change the order of computed values in a way corresponding to the sorted suffixes.
The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
    \caption{
        Segment IDs and segment positions of a segment join. Permuting these
        according to the inverse permutation of a suffix array results in 
        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
    }
    \label{fig:ids_and_positions}
\end{figure}

\input{sa_table.tex}

In the Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in a second
table \ref{tab:segment}.
For each segment, we store it's length, it's starting positions in the pangenome
and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts we use
a path join.
Similarly to segment join, a path join is a concatenation with delimiters $\#$
and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in the Figure \ref{fig:path_join}.
Starting positions can then be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.
Computation of the ranks is more involved.
We need to construct the suffix array of the path join and find it's inverse
permutation $ISA$.
This gives us ranks for each position in the path join.
To determine the rank of the right context for position $i$, we simply take
the value of the ranks $ISA[i+1]$.
Finaly, we store the starts and ranks sorted by the rank values in the segment
table as shown in Table \ref{tab:segment}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/path_join.png}
    \caption{}
    \label{fig:path_join}
\end{figure}

\input{segment_table.tex}

\subsection{Iterating suffix array}
With the previous tables in memory, we have all we need to generate the suffix
array entries.
Due to the prefix-freeness of the segments and the construction of the segment join
we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
Then, each row in the suffix table represents a suffix of a particular segment.

There are three cases where a letter at position $SA[i]$ can occur in these suffix segments:
\begin{itemize}
    \item it is a separator or the sentinel
    \item it is at the end of a segment inside the trigger part
    \item it is part of the segment outside the trigger part, separator and sentinel
\end{itemize}

In the first case, there is no corresponding position in the original pangenome,
therefore we can skip these rows of the suffix table.
% We can skip the first $n+1$ rows of the suffix table, because these suffixes
% do not yield any suffix array entry.

In the second case, the position is inside the trigger string.
Since the possitions inside the triggers are represented twice, once at the end
of a segment a second time at the beginning of the following segment, we need
to choose when to report them.
A natural choice is to report positions at the beginning of a segment and ignore
positions at the end.
Here we note that this choice also plays nicely with the previous choice of sentinel
characters, since ignoring last $k$ characters will result in not printing
sentinels either.
In conclusion, if the length of a current segment suffix is smaller or equal to
the size of the trigger words, we skip the row similarly as in the previous case.

Finaly, in the last case we report the positions.
The last case will partition the suffix table into the blocks of the same suffixes.
As an example consider the rows 20 and 21, which together for a single block.
All other blocks in the example consist of single rows and therefore we call them singletons.

This partitioning leads to three cases (from simplest to hardest):
\begin{itemize}
    \item A singleton block, occuring only once in the whole pangenome
    \item A singleton block, repeating several times in the pangenome
    \item A non-singleton block
\end{itemize}

In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
Given the row index $i$, this entry can be calculated with the following equation:

\begin{equation}
    \label{eq:sa_entry}
    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
\end{equation}

As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.

The second, slightly more complex case is a singleton block with several occurences in the pangenome.
For this case, we need to report as many suffix array entries as the number of occurences.
Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.

As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.

In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
These suffixes represent the same substring of the original pangenome.
Here, we report a suffix array entry for each occurence of the substring in the pangenome.
To identify which entry is the first, we need to find the starting position with the smallest right context rank.
Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.

As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.

\subsection{Application}
There are several proposed data structures implicitly relying on the structure
of prefix-free graphs.

Rpair: rescaling repair with rsync \cite{2019gagie}
Prefix-free parsing for building big BWTs \cite{2019boucher}
MONI and r-index \cite{rossi2022moni,gagie2020fully}
Prefix-free Wheeler graphs \cite{2022pfwg}
MARIA \cite{2022maria}

All of these works implement their own version of prefix-free graphs.
We argue that separating prefix-free graphs as a standalone data structure can
bring several benefits:

\begin{itemize}
    \item reduce the complexity of the presentation
    \item allowing for better optimization of algorithms related to prefix-free graphs
    \item supporting the theoretical research by clearly deliminating the relevant terms
    \item allowing to bring prefix-free graphs closer to the biological data
\end{itemize}

Related to the last point, we remark that a possible future research can be in
the proper choice of trigger words.
As an example, we used stop codons in our experiments.
This could create prefix-free graphs capturing some biologically relevant phenomena.
Other options are recombination hotspots, highly repetitive elements, different
binding sites and sites with increased breakage.

### 2023-06-22 manuscript/introduction.tex
finished intro

@@
 \section{Introduction}
-Suffix array is a data structure used in many fields, such as stringology,
-bioinformatics, and... .
-Particularly in bioinformatics, it is used as part of FM index, which in turn
-is used for read mapping.
-Recently, a new paradigm in bioinformatics was formed, moving from a single
-reference genome to a pangenome.
-The term pangenome was first used by \citet{tettelin2005genome} in 2005 during the
-resequencing experiments on streptococcus.
-Since then, the definition of a pangenome has shifted to any set of sequences
-analysed jointly \cite{computational2018computational}.
-Nevertheless, most pangenomes consist of highly related sequences and thus are
-highly repetitive.
-This leads to representation of pangenomes as graphs, where similar genome
-regions are represented by a single node and the nodes are conected to paths
-representing the original sequences.
-Several approaches of pangenomic graph construction exist, such as variation
-graphs \cite{church2015extending}?, cactus graphs \cite{paten2011cactus}, 
-Wheeler graphs \cite{gagie2017wheeler} ...
-Most of these approaches require an initial local alignment of similar regions,
-which makes them computationally expensive.
-Here we present a new class of graphs, prefix-free graphs, which are easy to
-construct and provide a way how to efficiently index them to allow pattern 
-search queries.
-Pangenomes introduced by...
-In computational pangenomics...
-Repetitive datasets...
-Graph pangenomes and textual pangenomes...
-Prefix-free parsing is a technique...
-Prefix-free graphs play an important role in improving space
-complexity of several efficient algorithms, but are never the concept is never
-framed as a pangenomic data structure.
-We think that separating and clearly defining this data structure can lead to
-more advancements...
-Our contribution is in defining a data structure called prefix-free graphs and
-showing how this data structure can be used to generate suffix array in
-sublinear space.
-Furthermore, we implemented two binaries for an easy construction of prefix-free
-graphs from a set of sequences, and from a graph in gfa format.
-Furthermore, we implemented rust library for working with prefix-free graphs.
-This library contains an iterator, which can be directly used to generate the 
-suffix array in sublinear space.
+The term pangenome was first used by \citet{tettelin2005genome} in 2005 while studying variations in the population of Streptococcus agalactiae.
+Since then, pangenomes have found applications in the study of many organisms, from viruses \cite{lau2021profiling} through microbes \cite{dutilh2014comparative} and plants \cite{danilevicz2020plant} to humans \cite{wang2022human}.
+As per the definition by The Computational Pan-Genomics Consortium \cite{computational2018computational}, a pangenome is any set of genomic sequences meant to be analyzed jointly.
+Nevertheless, in practice, most pangenomes consist of genomic sequences of highly related organisms and therefore are highly repetitive.
+Representation of this repetitive dataset by simple text is often inefficient and limits scaling in terms of algorithmic time and space complexity.
+These limitations lead to the idea of representing pangenomes as graphs, where similar genomic regions are unified into nodes, and these nodes are connected to paths representing the original genomic sequences.
+
+Several approaches to pangenomic graph construction exist, such as variation graphs \cite{church2015extending,garrison2018variation,garrison2023building}, cactus graphs \cite{paten2011cactus,hickey2023pangenome}, and Wheeler graphs \cite{gagie2017wheeler,2022pfwg}.
+Most of these approaches require an initial local alignment of similar regions or a multiple sequence alignment, which makes them computationally expensive.
+Here we present a new class of graphs, prefix-free graphs, which are orders of magnitude faster to construct.
+Furthermore, we explore the connection between prefix-free graphs and suffix arrays.
+
+A suffix array is a data structure from the stringology field with a massive impact on designing many efficient algorithms on strings.
+Particularly in bioinformatics, it is responsible for the design of such data structures as the Burrows-Wheeler transform \cite{burrows1994block} or FM-index \cite{ferragina2000opportunistic}, which in turn allowed for efficient mapping of reads to the reference and several other fundamental bioinformatics operations.
+These fundamental operations are well-studied on sequential data, but the recent paradigm shift of moving from a single reference genome to a graph pangenome made it even more complicated to apply the acquired knowledge from the stringology field to biological sequences.
+
+Thanks to the link to suffix arrays, prefix-free graphs have great potential to draw from this extensive knowledge.
+Using the suffix array from a prefix-free graph, we can obtain several stringological data structures which are not easily obtainable for graphs.
+This feature of prefix-free graphs was demonstrated in several articles \cite{2019boucher,rossi2022moni,2022pfwg,2022maria}, where the authors implicitly used similar techniques as presented here.
+We think that explicitly defining and framing the prefix-free graph as a standalone pangenomic data structure can bring several benefits:
+
+\begin{itemize}
+    \item reduction in the complexity of the presentation of several space-efficient algorithms
+    \item support of theoretical research by clearly delimiting the relevant terms
+    \item improved focus on the optimization of algorithms related to prefix-free graphs
+    \item enabling bringing prefix-free graphs closer to the biological data
+\end{itemize}
+
+In this work, we define prefix-free graphs and show how they can be constructed from the pangenome in its textual representation.
+Furthermore, we show in detail how prefix-free graphs can be used to generate the suffix array of a pangenome in sublinear space and linear time.
+Finally, we implement the presented algorithms as two binaries for easy construction of prefix-free graphs from a set of sequences in FASTA format and from a pangenomic graph in GFA format.
+Furthermore, we implement the rust library for working with prefix-free graphs.
+This library contains an iterator, which can be directly used to generate the suffix array in sublinear space.

### 2023-06-23 manuscript/pfg.tex
finished pfg

@@
 \section{Prefix-free graphs}
-% some intro?
-The idea of prefix-free graphs is inspired by rsync \cite{kornblum2006identifying}.
-%% Context-triggered piecewise hashing (CTPH) is a technique for parsing
-%% strings into blocks such that long repeated substrings are parsed the same
-%% way (except possibly at the beginning or end of the substrings).
+The idea of prefix-free graphs is inspired by a technique used in the tool rsync named Context-Triggered Piecewise Hashing (CTPH) \cite{kornblum2006identifying}.
+CTPH uses a rolling hash to partition a string into substrings such that long repeated substrings are partitioned the same way.
+These substrings are then hashed with a traditional hash function and stored as a string signature.
+The signatures of several files are then compared to determine their changes.
 
-% what are prefix-free graphs
-Given a set of sequences representing a pangenome, we can partition each
-sequence into \emph{segment}s.
-These segments form nodes of the prefix-free graph.
-The edges of a graph represent the adjacencies of the segment pairs, which are 
-present in the original strings.
-The adjacent segments are overlapping by a fixed number of characters $k$.
-We will call these $k$-sized overlaps triggers.
-Another important feature of the segments is that they contain triggers only at
-the start or at the end of a segment.
-This feature guarantees that the set of segments is prefix-free, i.e. no
-segment is a prefix of another segment.
+In prefix-free graphs, we partition each sequence in a given pangenome into \emph{segment}s.
+These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
+The sequences are then represented as \emph{path}s in the graph.
+
+The segments have two essential characteristics, which makes them a good choice for nodes of a pangenomic graph.
+Similarly to CTPH, long repeated sequences will be partitioned the same way.
+Furthermore, no segment is a prefix of another, making a set of segments prefix-free.
+The second characteristic is crucial for connecting prefix-free graphs and suffix arrays, as will be presented in the next section.
+
+To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger is a string of length $k$.
+For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
+Then, for each sequence in pangenome $P$, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
+Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
+If the segment's sequence was not yet observed during the scan, we add it to the set of segments and assign a unique ID.
+Each time we append the corresponding ID to the path representing the original sequence.
+
+Two special cases happen during the sequence scan, one at the beginning, when no previous trigger word was encountered, and another at the end, when the last $k$ characters are sentinels.
+These are addressed by simply starting the first segment at the start of a sequence and ending the last at the sequence end.
+
+Notably, the adjacent segments overlap by exactly $k$ characters.
+Furthermore, trigger words occur only at the beginnings or ends of segments because any occurrence of a trigger word in the middle of a segment would break it into two.
+This feature and the choice of sentinels outside the sequence's alphabet guarantee that the set of segments is prefix-free.
+A sketch of the proof is shown in Figure \ref{fig:proof}.
 
@@
 
-% how to construct prefix free graphs
-To illustrate how to partition the sequences into segments consider a set of
-strings $S = {TACTACGTACT, TCGTACTACT, TCGTCGTACTACT}$.
-Next, assume we will partition the sequences using a set of trigger words 
-$T = {AC, CG}$.
+After the previous steps, the set of segments and the list of paths already represent a prefix-free graph.
+However, to simplify the usage of prefix-free graphs, we recommend normalizing them.
+During the normalization, we sort segments lexicographically and change their IDs to correspond to the lexicographical ranks.
+This relabeling is then also propagated to paths accordingly.
 
-In stringology, strings are often flanked by so called \emph{sentinel} = $\$$.
-In our case, we need to append $k$ sentinels to the sequences.
-The reasons for doing so will become clear in the suffix array construction section.
-
-Then, using a sliding window we can scan through the sequences and each time we
-encounter a trigger word, we create a new segment from the start of the previous
-trigger word to the end of the current trigger word.
-Two special cases happen at the beginning of a sequence, when no previous
-trigger was encountered, and at the end.
-In the former case, the segment starts simply at the start of a sequence and in
-the later case the segment ends at a sequence end.
-
-During the sequence scan, each unique segment is assigned an ID.
-The original sequence can then be represented as a list of IDs, also called 
-\emph{path}s.
-
-The result of this procedure is a set of segments and a list of paths.
-In the previous example, the segments are ${ TAC, ACTAC, ACG, CGTAC, ACT\$\$, TCG, CGTCG }$
-and the paths are $[  ]$.
-
-Because each segment is represented in a set $S$ only once, the sum of lengths
-of segments is usually much smaller than the length of original sequences.
-
-This set of segments, together with the list of paths are already a prefix-free
-graph, but in order to be useful for the next part, we need to normalize it.
-By normalizing, we mean to sort the segments lexicographically and relabel the
-paths accordingly.
-This will give us a list of segments and a list of paths which can be directly
-represented in the GFA format as shown in the Figure \ref{fig:gfa}.
+To illustrate the entire procedure, consider a set of sequences $P = {}$ and a set of trigger words $T = {AC, CG}$.
+After the partitioning, we obtain a set of segments with IDs ${}$ and a list of paths $[]$.
+After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.
 
@@
     \centering
-    % \includegraphics[width=\linewidth]{images/pfg.drawio.png}
-    % \includegraphics[width=0.5\linewidth]{images/pfg_gfa.v2.png}
     \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
@@
 
-% add graph figure
-
-The reconstruction of the original sequence can then be done by expanding 
-the particular path, ignoring last $k$ characters of each segment.
+From this representation, original sequences of a pangenome can be reconstructed by expanding the segment IDs in a particular path, ignoring the last $k$ characters of each segment.

### 2023-06-23 manuscript/pfg.tex
finished pfg

@@
 These substrings are then hashed with a traditional hash function and stored as a string signature.
-The signatures of several files are then compared to determine their changes.
+The signatures of several files are then compared to determine changes.
 
@@
 These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
-The sequences are then represented as \emph{path}s in the graph.
+The sequences are represented as \emph{path}s in the graph.
 
-The segments have two essential characteristics, which makes them a good choice for nodes of a pangenomic graph.
+The segments have two essential characteristics making them a good choice for nodes of a pangenomic graph.
 Similarly to CTPH, long repeated sequences will be partitioned the same way.
@@
 
-To illustrate the entire procedure, consider a set of sequences $P = {}$ and a set of trigger words $T = {AC, CG}$.
-After the partitioning, we obtain a set of segments with IDs ${}$ and a list of paths $[]$.
+To illustrate the entire procedure, consider a set of sequences $P = \{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
+After the partitioning, we obtain a set of segments with IDs $\{$\texttt{0:CAC}, \texttt{1:ACG}, \texttt{2:CGTAC}, \texttt{3:ACT..}, \texttt{4:ACAC}, \texttt{5:CGAC}$\}$ and a list of paths \texttt{[[0,1,2,3], [0,4,3], [0,1,5,3]]}.
 After the normalization, we get a prefix-free graph which can be directly represented in GFA format as shown in Figure \ref{fig:gfa}.

### 2023-06-23 manuscript/sa_construction.tex
finished pfg
finished iterator preparation

@@
 \section{Suffix array construction}
-Suffix array is a permutation of string positions which lexicographically
+A suffix array is a permutation of string positions which lexicographically
 sorts the suffixes of the string starting at that position.
-It is extremely powerful data structure with many applications in efficient
-algorithms on strings.
-Enhanced by additional data structures, such as for example longest common
-prefix (LCP) array, it can fully substitute suffix tree and solve problems,
-such as exact pattern matching, repeat finding, maximum exact matching (MEM)
-finding, document retrieval and many more.
+It is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.
 
-There exists several algorithms for suffix array construction in linear time
-\cite{} with several practical implementations \cite{}.
-Altough their time complexity is linear, these algorithms become bottlenecks in
-some application due to their linear space complexity.
-This is especially relevant in pangenomics, where the datasets often do not fit
-in the memory.
-Nonetheless, as we shown in previous sections, prefix-free graphs can represent
-the pangenome in substantialy smaller space and therefore can fit in computer
-memory.
+There exist several algorithms for suffix array construction in linear time \cite{skew2003karkkainen,sais2009nong,saoverview2007puglisi} with several practical implementations \cite{louza2017inducing,libdivsufsort}.
+Despite their linear time complexity, these algorithms become bottlenecks in some applications because of their linear space complexity.
+This observation is especially relevant in pangenomics, where the datasets often do not fit in the computer memory.
 
-Here we show another crucial advantage of prefix-free graphs - their ability to
-output the suffix array of the original string iterativelly.
-This allows us to generate the suffix array one-by-one with the possibility
-to compress or use in subsequential computations directly, giving us the ability
-to use the rich theory of suffix arrays without ever storing them.
+Here, we show another crucial advantage of prefix-free graphs.
+Although they do not offer any improvement in theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using them directly in subsequential computation or storing them in compressed form.
+This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.
 
 \subsection{Iterator preparation}
-In order to prepare the prefix-free graph to generate the suffix array of
-original pangenome we need to create several data structures.
-First, we concatenate all the prefix-free segments into a single string using
-a separator $\#$ and append a sentinel $\$$.
+To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
+First, we concatenate all the segments into a single string using a separator $\#$ and append a sentinel $\$$.
 We will call this concatenation \emph{segment join}.
-An example of a segment join is in the Figure \ref{fig:ids_and_positions}.
+An example of a segment join is in Figure \ref{fig:ids_and_positions}.
 
-Next, for the segment join we calculate the suffix array and the LCP array.
-For this we can use the linear time, linear space algorithms mentioned previously,
-since the segment join is usually much smaller than the original pangenome.
+Next, we calculate the segment join's suffix array and the longest common prefix (LCP) array.
+For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
+We note that these linear complexities are proportional to the length of the segment join, which is usually much smaller than the original pangenome.
 
-Next, for each suffix of a segment join, we need to calculate the corresponding
-segment ID and the segment position.
-These two arrays represent the current suffix of a segment join starts with a
-particular segment and the position in this segment.
-This can be computed using an inverse permutation of a suffix array of a segment
-join in linear time.
-To illustrate the procedure, consider the segment join of our running example from
-Figure \ref{fig:ids_and_positions}.
-Each position of the join can be assigned a segment ID and a position in current
-segment by linearly iterating the segment join and incrementing ID and position
-accordingly.
-Then, applying the inverse permutation of a suffix array to these arrays will
-change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
+Next, for each suffix of the segment join, we need to calculate the corresponding segment ID and position values.
+The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
+These arrays can be computed using an inverse permutation of a suffix array of a segment join in linear time.
+To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
+Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
+Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
+Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
 segment ID array and segment position array.
@@
     \caption{
-        Segment IDs and segment positions of a segment join. Permuting these
-        according to the inverse permutation of a suffix array results in 
-        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
+        Segment IDs and segment positions of a segment join. 
+        Permuting these according to the inverse permutation of a suffix array results in segment ID array and segment position array shown in Table \ref{tab:suffix}.
     }
@@
 
-In the Table \ref{tab:suffix}, one row can represent multiple positions of the
+In Table \ref{tab:suffix}, one row can represent multiple positions of the
 pangenome.
-To identify these positions, we store some additional information in a second
-table \ref{tab:segment}.
-For each segment, we store it's length, it's starting positions in the pangenome
-and the rank of the right context of these positions.
-To calculate the starting positions and the ranks of the right contexts we use
-a path join.
-Similarly to segment join, a path join is a concatenation with delimiters $\#$
-and sentinel $\$$, but now constructed by concatenating the paths.
-An example of a path join for our running example is in the Figure \ref{fig:path_join}.
-Starting positions can then be calculated by cumulatively summing the lengths
+To identify these positions, we store some additional information in Table \ref{tab:segment}.
+For each segment, we store its length, starting positions in the pangenome and the rank of the right context of these positions.
+To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
+Similarly to segment join, a path join is a concatenation with delimiters $\#$ and sentinel $\$$, but now constructed by concatenating the paths.
+An example of a path join for our running example is in Figure \ref{fig:path_join}.
+Starting positions can be calculated by cumulatively summing the lengths
 of the segments in path join and subtracting the overlaps.
-Computation of the ranks is more involved.
-We need to construct the suffix array of the path join and find it's inverse
-permutation $ISA$.
-This gives us ranks for each position in the path join.
-To determine the rank of the right context for position $i$, we simply take
-the value of the ranks $ISA[i+1]$.
-Finaly, we store the starts and ranks sorted by the rank values in the segment
-table as shown in Table \ref{tab:segment}.
+
+The computation of ranks is more involved.
+It uses the normalized form of prefix-free graphs since it relies on a lexicographically smaller ID in a path representing a lexicographically smaller segment.
+We construct the suffix array of the path join and find its inverse permutation $\texttt{ISA}$.
+$\texttt{ISA}$ gives us ranks for each position in the path join.
+To determine the rank of the right context for position $i$, we take
+the value of $\texttt{ISA}[i+1]$.
+Finally, we store the starts and ranks sorted by the rank values in the segment table as shown in Table \ref{tab:segment}.
 
@@
 Then, each row in the suffix table represents a suffix of a particular segment.
+
+% the prefix-free property hold also for all suffixes of segments longer than k
+% The size of prefix-free graphs can be 
+% Because each segment is represented in a set $S$ only once, the sum of lengths
+% of segments is usually much smaller than the length of original sequences.
 

### 2023-06-23 manuscript/sa_construction.tex
started sa construction

@@
 
-\subsection{Iterating suffix array}
-With the previous tables in memory, we have all we need to generate the suffix
-array entries.
-Due to the prefix-freeness of the segments and the construction of the segment join
-we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
-Then, each row in the suffix table represents a suffix of a particular segment.
+\subsection{Iteration}
+With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array entries.
 
-% the prefix-free property hold also for all suffixes of segments longer than k
-% The size of prefix-free graphs can be 
-% Because each segment is represented in a set $S$ only once, the sum of lengths
-% of segments is usually much smaller than the length of original sequences.
-
-There are three cases where a letter at position $SA[i]$ can occur in these suffix segments:
+Each row in the suffix table represents a suffix of a particular segment.
+There are four cases of what the first position of these suffixes can represent within the segments:
 \begin{itemize}
-    \item it is a separator or the sentinel
-    \item it is at the end of a segment inside the trigger part
-    \item it is part of the segment outside the trigger part, separator and sentinel
+    \item the sentinel
+    \item a separator
+    \item a position within the last $k$ characters of a segment
+    \item a position outside the last $k$ characters, separator and sentinel
 \end{itemize}
 
-In the first case, there is no corresponding position in the original pangenome,
-therefore we can skip these rows of the suffix table.
-% We can skip the first $n+1$ rows of the suffix table, because these suffixes
-% do not yield any suffix array entry.
+Since the original pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.
 
-In the second case, the position is inside the trigger string.
-Since the possitions inside the triggers are represented twice, once at the end
-of a segment a second time at the beginning of the following segment, we need
-to choose when to report them.
-A natural choice is to report positions at the beginning of a segment and ignore
-positions at the end.
-Here we note that this choice also plays nicely with the previous choice of sentinel
-characters, since ignoring last $k$ characters will result in not printing
-sentinels either.
-In conclusion, if the length of a current segment suffix is smaller or equal to
-the size of the trigger words, we skip the row similarly as in the previous case.
+In the third case, the position is inside the trigger word or the sentinels appended during the graph construction.
+The positions inside the trigger words are represented twice in the suffix table, once at the end of a segment and a second time at the beginning of the following segment in the pangenome.
+These ending positions can violate the prefix-free property of the segment suffixes and, therefore, can be sorted incorrectly.
+Skipping through these positions ensures the prefix-free property for the rest of the suffixes and also avoids double reporting.
+This choice also plays nicely with the previous choice of appending $k$ sentinels during the construction of a prefix-free graph, as these positions will not get reported either.
+In conclusion, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.
 
+% stoped here
 Finaly, in the last case we report the positions.

### 2023-06-24 manuscript/conclusion.tex
finished sa construction

--- 2023-06-22 manuscript/sa_construction.tex
@@
-\section{Suffix array construction}
-Suffix array is a permutation of string positions which lexicographically
-sorts the suffixes of the string starting at that position.
-It is extremely powerful data structure with many applications in efficient
-algorithms on strings.
-Enhanced by additional data structures, such as for example longest common
-prefix (LCP) array, it can fully substitute suffix tree and solve problems,
-such as exact pattern matching, repeat finding, maximum exact matching (MEM)
-finding, document retrieval and many more.
-
-There exists several algorithms for suffix array construction in linear time
-\cite{} with several practical implementations \cite{}.
-Altough their time complexity is linear, these algorithms become bottlenecks in
-some application due to their linear space complexity.
-This is especially relevant in pangenomics, where the datasets often do not fit
-in the memory.
-Nonetheless, as we shown in previous sections, prefix-free graphs can represent
-the pangenome in substantialy smaller space and therefore can fit in computer
-memory.
-
-Here we show another crucial advantage of prefix-free graphs - their ability to
-output the suffix array of the original string iterativelly.
-This allows us to generate the suffix array one-by-one with the possibility
-to compress or use in subsequential computations directly, giving us the ability
-to use the rich theory of suffix arrays without ever storing them.
-
-\subsection{Iterator preparation}
-In order to prepare the prefix-free graph to generate the suffix array of
-original pangenome we need to create several data structures.
-First, we concatenate all the prefix-free segments into a single string using
-a separator $\#$ and append a sentinel $\$$.
-We will call this concatenation \emph{segment join}.
-An example of a segment join is in the Figure \ref{fig:ids_and_positions}.
-
-Next, for the segment join we calculate the suffix array and the LCP array.
-For this we can use the linear time, linear space algorithms mentioned previously,
-since the segment join is usually much smaller than the original pangenome.
-
-Next, for each suffix of a segment join, we need to calculate the corresponding
-segment ID and the segment position.
-These two arrays represent the current suffix of a segment join starts with a
-particular segment and the position in this segment.
-This can be computed using an inverse permutation of a suffix array of a segment
-join in linear time.
-To illustrate the procedure, consider the segment join of our running example from
-Figure \ref{fig:ids_and_positions}.
-Each position of the join can be assigned a segment ID and a position in current
-segment by linearly iterating the segment join and incrementing ID and position
-accordingly.
-Then, applying the inverse permutation of a suffix array to these arrays will
-change the order of computed values in a way corresponding to the sorted suffixes.
-The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
-segment ID array and segment position array.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/ids_and_positions.png}
-    \caption{
-        Segment IDs and segment positions of a segment join. Permuting these
-        according to the inverse permutation of a suffix array results in 
-        segment ID array and segment position array shown in the Table \ref{fig:segment_sa}.
-    }
-    \label{fig:ids_and_positions}
-\end{figure}
-
-\input{sa_table.tex}
-
-In the Table \ref{tab:suffix}, one row can represent multiple positions of the
-pangenome.
-To identify these positions, we store some additional information in a second
-table \ref{tab:segment}.
-For each segment, we store it's length, it's starting positions in the pangenome
-and the rank of the right context of these positions.
-To calculate the starting positions and the ranks of the right contexts we use
-a path join.
-Similarly to segment join, a path join is a concatenation with delimiters $\#$
-and sentinel $\$$, but now constructed by concatenating the paths.
-An example of a path join for our running example is in the Figure \ref{fig:path_join}.
-Starting positions can then be calculated by cumulatively summing the lengths
-of the segments in path join and subtracting the overlaps.
-Computation of the ranks is more involved.
-We need to construct the suffix array of the path join and find it's inverse
-permutation $ISA$.
-This gives us ranks for each position in the path join.
-To determine the rank of the right context for position $i$, we simply take
-the value of the ranks $ISA[i+1]$.
-Finaly, we store the starts and ranks sorted by the rank values in the segment
-table as shown in Table \ref{tab:segment}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/path_join.png}
-    \caption{}
-    \label{fig:path_join}
-\end{figure}
-
-\input{segment_table.tex}
-
-\subsection{Iterating suffix array}
-With the previous tables in memory, we have all we need to generate the suffix
-array entries.
-Due to the prefix-freeness of the segments and the construction of the segment join
-we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
-Then, each row in the suffix table represents a suffix of a particular segment.
-
-There are three cases where a letter at position $SA[i]$ can occur in these suffix segments:
-\begin{itemize}
-    \item it is a separator or the sentinel
-    \item it is at the end of a segment inside the trigger part
-    \item it is part of the segment outside the trigger part, separator and sentinel
-\end{itemize}
-
-In the first case, there is no corresponding position in the original pangenome,
-therefore we can skip these rows of the suffix table.
-% We can skip the first $n+1$ rows of the suffix table, because these suffixes
-% do not yield any suffix array entry.
-
-In the second case, the position is inside the trigger string.
-Since the possitions inside the triggers are represented twice, once at the end
-of a segment a second time at the beginning of the following segment, we need
-to choose when to report them.
-A natural choice is to report positions at the beginning of a segment and ignore
-positions at the end.
-Here we note that this choice also plays nicely with the previous choice of sentinel
-characters, since ignoring last $k$ characters will result in not printing
-sentinels either.
-In conclusion, if the length of a current segment suffix is smaller or equal to
-the size of the trigger words, we skip the row similarly as in the previous case.
-
-Finaly, in the last case we report the positions.
-The last case will partition the suffix table into the blocks of the same suffixes.
-As an example consider the rows 20 and 21, which together for a single block.
-All other blocks in the example consist of single rows and therefore we call them singletons.
-
-This partitioning leads to three cases (from simplest to hardest):
-\begin{itemize}
-    \item A singleton block, occuring only once in the whole pangenome
-    \item A singleton block, repeating several times in the pangenome
-    \item A non-singleton block
-\end{itemize}
-
-In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
-Given the row index $i$, this entry can be calculated with the following equation:
-
-\begin{equation}
-    \label{eq:sa_entry}
-    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
-\end{equation}
-
-As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
-It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
-The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
-This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.
-
-The second, slightly more complex case is a singleton block with several occurences in the pangenome.
-For this case, we need to report as many suffix array entries as the number of occurences.
-Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.
-
-As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
-This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
-Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.
-
-In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
-These suffixes represent the same substring of the original pangenome.
-Here, we report a suffix array entry for each occurence of the substring in the pangenome.
-To identify which entry is the first, we need to find the starting position with the smallest right context rank.
-Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
-To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.
-
-As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
-The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
-The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
-Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.
-
+\section{Conclusion}
 \subsection{Application}

### 2023-06-24 manuscript/sa_construction.tex
finished sa construction

@@
 \subsection{Iteration}
-With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array entries.
+With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array values.
 
@@
 
-% stoped here
-Finaly, in the last case we report the positions.
-The last case will partition the suffix table into the blocks of the same suffixes.
-As an example consider the rows 20 and 21, which together for a single block.
-All other blocks in the example consist of single rows and therefore we call them singletons.
+Finally, in the last case, we report the suffix array values.
+The suffix table can be partitioned into blocks of the same segment suffixes.
+For example, consider rows 20 and 21, which form a single block.
+All other blocks in the running example consist of single rows; therefore, we call them singletons.
 
-This partitioning leads to three cases (from simplest to hardest):
+This partitioning leads to three cases:
 \begin{itemize}
-    \item A singleton block, occuring only once in the whole pangenome
-    \item A singleton block, repeating several times in the pangenome
-    \item A non-singleton block
+    \item a singleton block with segment suffix occurring only once in the whole pangenome
+    \item a singleton block with segment suffix occurring several times in the pangenome
+    \item a non-singleton block
 \end{itemize}
 
-In the case of a singleton block with only a single occurence we need to report only a single suffix array entry.
-Given the row index $i$, this entry can be calculated with the following equation:
+In the first case, we must report only a single suffix array value.
+Given the row index $i$, this value can be calculated with Equation \ref{eq:sa_value}.
 
 \begin{equation}
-    \label{eq:sa_entry}
-    \texttt{SA entry} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
+    \label{eq:sa_value}
+    \texttt{SA value} = \texttt{starts}[\texttt{ID}[i]] + \texttt{pos}[i]
 \end{equation}
@@
 As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
-It's segment ID is $0$ and from the Segment Table (\ref{tab:segment}) we see that there is only one occurence of segment $0$ and it starts at position $9$ in the pangenome.
+Its segment ID is $0$, and from the Segment Table (\ref{tab:segment}), we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
 The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
-This yields the first entry of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.
+Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.
 
-The second, slightly more complex case is a singleton block with several occurences in the pangenome.
-For this case, we need to report as many suffix array entries as the number of occurences.
-Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_entry} to each of them.
+The second, slightly more complex case is a singleton block representing a segment suffix with several occurrences in the pangenome.
+In this case, we must report as many suffix array values as the number of occurrences.
+Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.
 
@@
 This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
-Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array entry $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.
+Since the offset from the start of a segment $\texttt{pos}[14]$ is $0$, we report a suffix array values $15 + 0 = 15$ and $1 + 0 = 1$, corresponding to the suffixes $P[15..] = \texttt{ACGACT}$ and $P[1..] = \texttt{ACGTACT}$.
 
-In the last case, we have a non-singleton block with suffixes of several segments, possibly with multiple occurences.
-These suffixes represent the same substring of the original pangenome.
-Here, we report a suffix array entry for each occurence of the substring in the pangenome.
-To identify which entry is the first, we need to find the starting position with the smallest right context rank.
+In the last case, we have a non-singleton block representing suffixes of several segments, possibly with multiple occurrences.
+These suffixes represent identical substrings in the original pangenome.
+Here, we report a suffix array value for each of the substrings.
+To identify the first value, we must find the starting position with the smallest right context rank.
 Because the ranks are sorted, this procedure is similar to the merging phase of a merge sort.
-To iterate through all suffix array entries in the block, we therefore always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_entry} to this segment start.
+Therefore, to iterate through all suffix array values in the block, we always identify the segment start with the next smallest right context rank and apply Equation \ref{eq:sa_value} to this segment start.
 
-As an example, consider the block at rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
+As an example, consider the block of rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
 The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.
 The right context ranks from smallest to highest are $4, 5, 6, 9$ with the corresponding segment starts $8, 14, 0, 9$.
-Applying Equation \ref{eq:sa_entry} to these segment starts yields a suffix array entries $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.
-
-\subsection{Application}
-There are several proposed data structures implicitly relying on the structure
-of prefix-free graphs.
-
-Rpair: rescaling repair with rsync \cite{2019gagie}
-Prefix-free parsing for building big BWTs \cite{2019boucher}
-MONI and r-index \cite{rossi2022moni,gagie2020fully}
-Prefix-free Wheeler graphs \cite{2022pfwg}
-MARIA \cite{2022maria}
-
-All of these works implement their own version of prefix-free graphs.
-We argue that separating prefix-free graphs as a standalone data structure can
-bring several benefits:
-
-\begin{itemize}
-    \item reduce the complexity of the presentation
-    \item allowing for better optimization of algorithms related to prefix-free graphs
-    \item supporting the theoretical research by clearly deliminating the relevant terms
-    \item allowing to bring prefix-free graphs closer to the biological data
-\end{itemize}
-
-Related to the last point, we remark that a possible future research can be in
-the proper choice of trigger words.
-As an example, we used stop codons in our experiments.
-This could create prefix-free graphs capturing some biologically relevant phenomena.
-Other options are recombination hotspots, highly repetitive elements, different
-binding sites and sites with increased breakage.
+Applying Equation \ref{eq:sa_value} to these segment starts yields a suffix array values $8 + 0 = 8$, $14 + 0 = 14$, $0 + 0 = 0$ and $9 + 1 = 10$, representing pangenome suffixes $P[8..] = \texttt{CACACT}$, $P[14..] = \texttt{CACGACT}$, $P[0..] = \texttt{CACGTACT}$, and $P[10..] = \texttt{CACT}$.

### 2023-06-24 manuscript/results.tex
finished results

@@
 \section{Results}
-We implemented the prefix-free graphs package in rust programming language.
-There are two binary crates and one library crate included in the package.
-Binary crates serve the purpose of creating prefix-free graphs from FASTA and from GFA formats.
-Library crate provides an interface for working with the prefix-free graphs,
-mainly an iterator of suffix array in sublinear space and linear time.
+We implemented the prefix-free graphs as a Rust package.
+This package contains two binary crates and one library crate.
+Binary crates are executables that serve the purpose of creating prefix-free graphs from FASTA and GFA formats.
+The binary crates are named \texttt{fasta2pfg} and \texttt{gfa2pfg}, and their usage is as follows:
 
-The binary crates are fasta2pfg and gfa2pfg with usage:
 \begin{minted}{bash}
@@
 
-The library can be used from within rust programming language as follows:
+We compared the running time of the construction algorithm with several pangenome construction tools, namely with the PanGenome Graph Builder \cite{garrison2023building}, VG \cite{garrison2018variation}, and Minigraph-Cactus \cite{hickey2023pangenome}.
+For comparison, we used up to 256 SARS-CoV-2 sequences.
+These sequences have lengths of around 29.9 kbp and high nucleotide similarity.
+The resulting running times are shown in Table \ref{tab:comparison}.
+
+\input{comparison_table.tex}
+
+The library crate provides an interface for working with prefix-free graphs, mainly an iterator of a suffix array in sublinear space and linear time.
+The library can be used from within the Rust programming language as follows:
 
@@
 
-for (i, sa_i) in pfg.iter().enumerate() {
-    println!("{}\t{}", i, sa_i);
+for (i, (sa_i, id_i, pos_i)) in pfg.iter().enumerate() {
+    println!("{}\t{}\t{}\t{}", i, sa_i, id_i, pos_i);
 }
@@
 
-We compared the pangenome construction of several tools.
-
-\input{comparison_table.tex}
+In addition to the suffix array value, the iterator provides the segment ID and position values.
+These are useful for the computation of several related data structures.
+Similarly to \citet{2019boucher}, we can use them to compute the Burrows-Wheeler transform $\texttt{BWT}[i]$ by storing the preceding characters of each segment and then reporting the character at position $\texttt{pos}[i]-1$ of a segment $\texttt{id}[i]$.
+This computation allows for space-efficient construction of r-index \cite{gagie2020fully} similar to MONI implementation \cite{rossi2022moni}, space-efficient construction of Wheeler graphs similar in nature to the implementation in \citet{2022pfwg}, and with the use of predecessor queries, space-efficient construction of a tag array introduced in \citet{2022maria}.

### 2023-06-24 manuscript/conclusion.tex
added conclusion

@@
 \section{Conclusion}
-\subsection{Application}
-There are several proposed data structures implicitly relying on the structure
-of prefix-free graphs.
-
-Rpair: rescaling repair with rsync \cite{2019gagie}
-Prefix-free parsing for building big BWTs \cite{2019boucher}
-MONI and r-index \cite{rossi2022moni,gagie2020fully}
-Prefix-free Wheeler graphs \cite{2022pfwg}
-MARIA \cite{2022maria}
-
-All of these works implement their own version of prefix-free graphs.
-We argue that separating prefix-free graphs as a standalone data structure can
-bring several benefits:
-
-\begin{itemize}
-    \item reduce the complexity of the presentation
-    \item allowing for better optimization of algorithms related to prefix-free graphs
-    \item supporting the theoretical research by clearly deliminating the relevant terms
-    \item allowing to bring prefix-free graphs closer to the biological data
-\end{itemize}
-
-Related to the last point, we remark that a possible future research can be in
-the proper choice of trigger words.
-As an example, we used stop codons in our experiments.
-This could create prefix-free graphs capturing some biologically relevant phenomena.
-Other options are recombination hotspots, highly repetitive elements, different
-binding sites and sites with increased breakage.
+We introduced the prefix-free graph as a standalone data structure in pangenomics and showed how to use it to build a suffix array of a pangenome in sublinear space.
+There are several possible directions for future research on prefix-free graphs.
+The first direction may be to answer the question of which additional stringological algorithms and data structures can use prefix-free graphs to improve their applicability to such vast and repetitive datasets as pangenomic data.
+Next, the choice of trigger words is, so far, mostly undiscovered territory.
+From a computational perspective, the set of trigger words minimizing the size of a resulting prefix-free graph is desirable since it would allow the analysis of larger datasets.
+Furthermore, trigger words offer great flexibility in the construction process.
+We hypothesize this flexibility could be used for the construction of prefix-free graphs based on biologically significant strings.
+In our experiments, we used the stop codon sequences as the trigger words, but other motives, such as different binding sites, recombination hotspots, or repetitive elements, are possible and could better capture some biological phenomena.

### 2023-06-24 manuscript/main.tex
finished pfg
added abstract

@@
 \author[1]{Andrej Baláž}[
-    orcid=0000-0001-0002-0003,
     email=andrejbalaz001@gmail.com,
@@
 
-\author[1]{Alessia Petescia}[
-    orcid=0000,
-    email=alessia@gmail.com,
-]
+\author[1]{Alessia Petescia}[]
 
@@
 \begin{abstract}
-  Recently, many graph structures for computational pangenomics were proposed.
-  In order to be useful, these graph structures need to implement several
-  operations such as efficient construction from many complete genomes,
-  mapping of short and long read.
-  These basic bioinformatics operations are well studied on sequential data,
-  and together with data structures such as suffix trees, suffix arrays and 
-  burrows wheeler transform allow high performance.
-  Attempts to implement these operations on graphs brings with it many complications
-  since these date structures are not easily obtainable for graphs.
-  In this work, we introduce prefix-free graphs, a pangenomic data structure
-  allowing to obtain the well known data structures from stringology in graph
-  settings, which in turn allows for many efficient operations on pangenomes.
+    A recent paradigm shift in bioinformatics from a single reference genome to a pangenome brought with it several graph structures.
+    These graph structures must implement operations, such as efficient construction from multiple genomes and read mapping.
+    Read mapping is a well-studied problem in sequential data, and, together with data structures such as suffix array and Burrows-Wheeler transform, allows for efficient computation.
+    Attempts to achieve comparatively high performance on graphs bring many complications since the common data structures on strings are not easily obtainable for graphs.
+    In this work, we introduce prefix-free graphs, a pangenomic data structure allowing us to obtain the well-known data structures from stringology in graph settings, allowing for many efficient operations on pangenomes.
 \end{abstract}
@@
 \begin{keywords}
-  pangenome \sep
+    computational pangenomics \sep
+    graph pangenome \sep
     suffix array
@@
 \input{introduction.tex}
-
 \input{pfg.tex}
-
 \input{sa_construction.tex}
-
 \input{results.tex}
-
 \input{conclusion.tex}

### 2023-06-24 manuscript/main.tex
added description to figures

@@
 Attempts to achieve comparatively high performance on graphs bring many complications since the common data structures on strings are not easily obtainable for graphs.
-    In this work, we introduce prefix-free graphs, a pangenomic data structure allowing us to obtain the well-known data structures from stringology in graph settings, allowing for many efficient operations on pangenomes.
+In this work, we introduce prefix-free graphs, a novel pangenomic data structure; we show how to construct them and how to use them to obtain well-known data structures from stringology in sublinear space, allowing for many efficient operations on pangenomes.
 \end{abstract}

### 2023-06-24 manuscript/pfg.tex
started sa construction
added description to figures

@@
     \centering
-    \includegraphics[width=\linewidth]{images/prefixfree_proof.png}
-    \caption{Visual proof that the trigger words induce prefix-free segments.}
+    \includegraphics[width=0.6\linewidth]{images/prefixfree_proof.png}
+    \caption{
+        A sketch of a visual proof that the trigger words induce prefix-free segments.
+        Suppose a segment is a prefix of another segment.
+        By construction, it ends with a trigger word.
+        This trigger word would be in the middle of another segment and would break it into two smaller segments.
+    }
     \label{fig:proof}
@@
     \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
-    \caption{Prefix-free graph in GFA format. Link lines omitted for brevity.}
+    \caption{
+        Prefix-free graph of the running example after normalization and its representation in GFA format.
+        Link lines omitted for brevity.
+    }
     \label{fig:gfa}

### 2023-06-24 manuscript/sa_construction.tex
added description to figures

@@
     \caption{
-        Segment IDs and segment positions of a segment join. 
-        Permuting these according to the inverse permutation of a suffix array results in segment ID array and segment position array shown in Table \ref{tab:suffix}.
+        Segment join of segments from the running example, corresponding segment IDs, and positions.
+        Permuting IDs and positions according to the inverse permutation of the suffix array results in the segment ID array and the segment position array shown in Table \ref{tab:suffix}.
     }
@@
     \includegraphics[width=\linewidth]{images/path_join.png}
-    \caption{}
+    \caption{
+        Path join of paths from the running example and the corresponding starts and ranks. The pangenome and segments are shown above to clarify the meaning of the start values.
+    }
     \label{fig:path_join}

### 2023-06-24 manuscript/comparison_table.tex
finished iterator preparation
finished?

@@
 \begin{table*}
-    \label{tab:segment}
-    \begin{tabular}{rllll}
+    \label{tab:comparison}
+    \begin{tabular}{rrrrr}
         \toprule
-        \#sequences & pggb & vg & minigraph-cactus & PFG \\
+        \#seqs & pggb & vg & minigraph-cactus & pfg \\
         \midrule
-        % 2 &
-        4   & 0m12s & 0m1s  & 3m47s  & 0m1s \\
-        16  & 0m19s & 0m1s  & 6m30s  & 0m1s \\
-        64  & 1m20s & 0m2s  & 13m30s & 0m1s \\
-        256 & 7m20s & 0m55s & 41m49s & 0m3s \\
+        4   & 0m 12s & 0m 1s  & 3m 47s  & 0m 1s \\
+        16  & 0m 19s & 0m 1s  & 6m 30s  & 0m 1s \\
+        64  & 1m 20s & 0m 2s  & 13m 30s & 0m 1s \\
+        256 & 7m 20s & 0m 55s & 41m 49s & 0m 3s \\
         \bottomrule
@@
         Comparison of running times of pangenomic graph construction tools.
-        Some of these tools need some kind of preprocessing in order to build the graph.
-        pggb requires for the fasta files to be indexed.
-        In our case, due to the small size of our files, this time was negligible, but it can have a bigger impact when bigger genomes are considered.
-        VG builds a pangenome graph starting from a VCF file or a multiple sequence alignment.
-        In each case, those outputs require to perform time consuming steps.
-        Having a suitable VCF file requires to align each haplotype to a reference genome and subsequently performing variant calling.
-        Beyond being time consuming, using a linear reference in a first place can introduce some biases in the first place.
-        On the other hand, performing a multiple sequence alignment is also known to be potentially slow, according to the length and the abundance of the underlying sequences.
+        Wall clock times are measured by the built-in bash command \texttt{time} and rounded up to the nearest second.
+        In addition to the measured times, \texttt{pggb} and \texttt{vg} need preprocessing of the pangenome.
+        \texttt{pggb} requires the fasta file to be indexed.
+        This time was negligible in our case, but it can have a more significant impact when a more extensive dataset is considered.
+        \texttt{vg} builds a pangenome from a VCF file or a multiple sequence alignment.
+        In both cases, acquiring this data requires time-consuming steps.
+        Furthermore, variant calling using a linear reference can introduce a reference bias.
     }

### 2023-06-26 manuscript/conclusion.tex
finished

@@
 We hypothesize this flexibility could be used for the construction of prefix-free graphs based on biologically significant strings.
-In our experiments, we used the stop codon sequences as the trigger words, but other motives, such as different binding sites, recombination hotspots, or repetitive elements, are possible and could better capture some biological phenomena.
+In our experiments, we used the stop codon sequences as the trigger words, but other motives, such as different binding sites, recombination hotspots, or repetitive elements, are possible and could be used to capture particular biological phenomena.

### 2023-06-26 manuscript/main.tex
finished?
finished

@@
 
+\section{Online Resources}
+The sources for the prefix-free graphs are available via
+\begin{itemize}
+\item \href{https://github.com/andynet/pfg}{https://github.com/andynet/pfg}.
+\end{itemize}
+
 \begin{acknowledgments}
-    Thanks to Askar for his encouragement while developing prefix-free graphs.
-    Thanks to Travis for introducing us to the advanced stringology.
-    Thanks to Adrian for discussing some ideas from this paper with us.
+This research was funded by a grant from the European Union’s Horizon 2020 research and innovation programme under the Marie Skłodowska-Curie grant agreement No 956229 (ALPACA) and by a grant from Slovak Research Grant Agency VEGA 1/0538/22.
 \end{acknowledgments}
@@
 
-\appendix
-
-\section{Online Resources}
-
-The sources for the prefix-free graphs are available via
-\begin{itemize}
-\item \href{https://github.com/andynet/pfg}{GitHub}.
-\end{itemize}
-
 \end{document}

### 2023-06-26 manuscript/pfg.tex
finished

@@
 
-In prefix-free graphs, we partition each sequence in a given pangenome into \emph{segment}s.
+In prefix-free graphs, we partition each sequence of a given pangenome into \emph{segment}s.
 These segments form nodes of the prefix-free graph, and their adjacencies in the original sequences constitute edges.
@@
 
-To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger is a string of length $k$.
+To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
 For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.

### 2023-06-26 manuscript/sa_construction.tex
finished

@@
 Here, we show another crucial advantage of prefix-free graphs.
-Although they do not offer any improvement in theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using them directly in subsequential computation or storing them in compressed form.
+Although they do not offer any improvement of theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using the values directly in subsequential computation or storing them in compressed form.
 This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.
@@
 To prepare the iterator of a suffix array of the pangenome from a prefix-free graph, we need to create several data structures.
-First, we concatenate all the segments into a single string using a separator $\#$ and append a sentinel $\$$.
+First, we concatenate all the segments into a single string using a separator $\texttt{\#}$ and append a sentinel $\texttt{\$}$.
 We will call this concatenation \emph{segment join}.
@@
 The value segment ID represents in what segment the current suffix starts, and the value segment position represents at what position in that particular segment the current suffix starts.
-These arrays can be computed using an inverse permutation of a suffix array of a segment join in linear time.
+These arrays can be computed using an inverse permutation of a suffix array $\texttt{ISA}$ (Equation \ref{eq:isa}) of a segment join in linear time.
+
+\begin{equation}
+    \label{eq:isa}
+    \texttt{ISA}[\texttt{SA}[i]] = i
+\end{equation}
+
 To illustrate the procedure, consider the segment join of our running example from Figure \ref{fig:ids_and_positions}.
@@
 Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
-Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
-segment ID array and segment position array.
+The resulting suffix array, LCP array, segment ID array and segment positions array are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.
 
@@
 
-In Table \ref{tab:suffix}, one row can represent multiple positions of the
-pangenome.
-To identify these positions, we store some additional information in Table \ref{tab:segment}.
-For each segment, we store its length, starting positions in the pangenome and the rank of the right context of these positions.
+\input{segment_table.tex}
+
+In the suffix table, one row can represent multiple positions of the pangenome.
+To identify these positions, we store some additional information in a \emph{segment table}.
+For each segment, we store its length, starting positions in the pangenome and ranks of the right contexts of these positions.
 To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
-Similarly to segment join, a path join is a concatenation with delimiters $\#$ and sentinel $\$$, but now constructed by concatenating the paths.
+Similarly to segment join, a path join is a concatenation with delimiters $\texttt{\#}$ and sentinel $\texttt{\$}$, but now constructed by concatenating the paths.
 An example of a path join for our running example is in Figure \ref{fig:path_join}.
-Starting positions can be calculated by cumulatively summing the lengths
+Then, starting positions can be calculated by cumulatively summing the lengths
 of the segments in path join and subtracting the overlaps.
@@
 
-\input{segment_table.tex}
-
 \subsection{Iteration}
@@
 Skipping through these positions ensures the prefix-free property for the rest of the suffixes and also avoids double reporting.
+Therefore, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.
 This choice also plays nicely with the previous choice of appending $k$ sentinels during the construction of a prefix-free graph, as these positions will not get reported either.
-In conclusion, if the length of a current segment suffix is smaller or equal to the size of the trigger words $k$, we skip the row as in the previous cases.
 
@@
 
-As an example, consider the row $13$ in the Suffix Table (\ref{tab:suffix}), the first row yielding a SA value.
-Its segment ID is $0$, and from the Segment Table (\ref{tab:segment}), we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
+As an example, consider the row $13$ in Table \ref{tab:suffix}, the first row yielding a SA value.
+Its segment ID is $0$, and from Table \ref{tab:segment}, we see only one occurrence of segment $0$ with starting position $9$ in the pangenome.
 The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
@@
 In this case, we must report as many suffix array values as the number of occurrences.
-Because the starting segment positions in the Segment Table (\ref{tab:segment}) are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.
+Because the starting segment positions in the segment table are sorted based on their right context rank, we can iterate through these starting positions and apply Equation \ref{eq:sa_value} to each of them.
 
-As an example, consider the row $14$ in the Suffix Table (\ref{tab:suffix}).
+As an example, consider the row $14$ in Table \ref{tab:suffix}.
 This suffix occurs twice in the pangenome in segments starting at positions $15$ and $1$.
@@
 
-As an example, consider the block of rows $[20..21]$ in the Suffix Table (\ref{tab:suffix}).
+As an example, consider the block of rows $[20..21]$ in Table \ref{tab:suffix}.
 The relevant segment IDs are $0$ and $3$, with segment starts at positions $9$, $8$, $14$ and $0$.

### 2023-06-26 src/pf.rs
cleaned repository

@@ -75,7 +75,7 @@ impl PFData {
         let gfa = parser.parse_file(filename)
             .expect("Error parsing GFA file.");
 
-        let overlap = 1; // TODO determine_overlap();
+        let overlap = 2; // TODO determine_overlap();
         let segments: Vec<Vec<u8>> = parse_segments(&gfa.segments);
         let paths: Vec<Vec<usize>> = parse_paths(&gfa.paths);
 

### 2023-06-26 manuscript/content/01_introduction.tex
moved content to content

--- 2023-06-22 manuscript/introduction.tex
@@
 A suffix array is a data structure from the stringology field with a massive impact on designing many efficient algorithms on strings.
-Particularly in bioinformatics, it is responsible for the design of such data structures as the Burrows-Wheeler transform \cite{burrows1994block} or FM-index \cite{ferragina2000opportunistic}, which in turn allowed for efficient mapping of reads to the reference and several other fundamental bioinformatics operations.
+Particularly in bioinformatics, it is responsible for the design of such data structures as the Burrows-Wheeler transform \cite{burrows1994block} and FM-index \cite{ferragina2000opportunistic}, which in turn allowed for efficient mapping of reads to the reference and several other fundamental bioinformatics operations.
 These fundamental operations are well-studied on sequential data, but the recent paradigm shift of moving from a single reference genome to a graph pangenome made it even more complicated to apply the acquired knowledge from the stringology field to biological sequences.

### 2023-08-09 manuscript/content/01_introduction.tex
integrated reviewers' suggestions

@@
 Using the suffix array from a prefix-free graph, we can obtain several stringological data structures which are not easily obtainable for graphs.
-This feature of prefix-free graphs was demonstrated in several articles \cite{2019boucher,rossi2022moni,2022pfwg,2022maria}, where the authors implicitly used similar techniques as presented here.
+This feature of prefix-free graphs was implicitly demonstrated in several articles \cite{2019boucher,rossi2022moni,2022pfwg,2022maria}, where the authors used similar techniques as presented here.
 We think that explicitly defining and framing the prefix-free graph as a standalone pangenomic data structure can bring several benefits:

### 2023-08-09 manuscript/content/02_pfg.tex
removed unnecessary images
integrated reviewers' suggestions

--- 2023-06-26 manuscript/pfg.tex
@@
 
-To create a prefix-free graph from a given pangenome $P$, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
+To create a prefix-free graph from a given pangenome, we define a set of trigger words $T$, where each trigger word is a string of length $k$.
 For this set $T$, we build an Aho-Corasick automaton \cite{aho1975efficient}.
-Then, for each sequence in pangenome $P$, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
+Then, for each sequence in the pangenome, we append $k$ sentinel characters and iterate over such modified sequence, searching for matches with set $T$ using the automaton.
 Each time we encounter a trigger word, we recognize a new segment from the start of the previous trigger word to the end of the current trigger word.
@@
 This relabeling is then also propagated to paths accordingly.
+All construction steps can be performed in space proportional to the sum of segment lengths and the sum of path lengths, which is expected to be significantly smaller than the length of the pangenome.
 
-To illustrate the entire procedure, consider a set of sequences $P = \{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
+To illustrate the entire procedure, consider a set of sequences $\{$\texttt{CACGTACT}, \texttt{CACACT}, \texttt{CACGACT}$\}$ and a set of trigger words $T = \{$\texttt{AC}, \texttt{CG}$\}$.
 After the partitioning, we obtain a set of segments with IDs $\{$\texttt{0:CAC}, \texttt{1:ACG}, \texttt{2:CGTAC}, \texttt{3:ACT..}, \texttt{4:ACAC}, \texttt{5:CGAC}$\}$ and a list of paths \texttt{[[0,1,2,3], [0,4,3], [0,1,5,3]]}.
@@
     \centering
-    \includegraphics[width=\linewidth]{images/pfg_gfa.v3.png}
+    \includegraphics[width=\linewidth]{images/pfg_gfa.png}
     \caption{

### 2023-08-09 manuscript/content/03_sa_construction.tex
moved content to content
integrated reviewers' suggestions

--- 2023-06-26 manuscript/sa_construction.tex
@@
 \section{Suffix array construction}
-A suffix array is a permutation of string positions which lexicographically
-sorts the suffixes of the string starting at that position.
+A suffix array is a permutation of string positions which lexicographically sorts the suffixes of the string starting at that position.
 It is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.
@@
 
-Next, we calculate the segment join's suffix array and the longest common prefix (LCP) array.
+Next, we calculate the segment join's suffix array and the longest common prefix array \cite{landau2001linear,manzini2004two}.
 For both of these arrays, there exist algorithms with linear time and space complexity which we can use.
@@
 Each position of the join can be assigned a segment ID and a position in the current segment by linearly scanning the segment join and incrementing the ID and position accordingly.
-Then, applying the inverse permutation of a suffix array to these arrays changes the order of computed values in correspondence to the sorted suffixes.
-The resulting suffix array, LCP array, segment ID array and segment positions array are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.
+Then, applying the ISA to these arrays changes the order of computed values in correspondence to the sorted suffixes.
+The resulting suffix array $\texttt{SA}$, longest common prefix array $\texttt{LCP}$, segment ID array $\texttt{ID}$ and segment positions array $\texttt{pos}$ are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.
 
@@
     Segment join of segments from the running example, corresponding segment IDs, and positions.
-        Permuting IDs and positions according to the inverse permutation of the suffix array results in the segment ID array and the segment position array shown in Table \ref{tab:suffix}.
+Permuting IDs and positions according to the \texttt{ISA} results in the segment ID array \texttt{ID} and the segment position array \texttt{pos} shown in Table \ref{tab:suffix}.
+\label{fig:ids_and_positions}
     }
-    \label{fig:ids_and_positions}
 \end{figure}
 
-\input{sa_table.tex}
+\input{content/table_suffix.tex}
 
-\input{segment_table.tex}
+\input{content/table_segment.tex}
 
 In the suffix table, one row can represent multiple positions of the pangenome.
+% These positions can be represented as a single integer coresponding to index in the concatenated representation of a pangenome.
 To identify these positions, we store some additional information in a \emph{segment table}.
-For each segment, we store its length, starting positions in the pangenome and ranks of the right contexts of these positions.
+For each segment of the prefix-free graph, we store its length, starting positions in the concatenated pangenome and ranks of the right contexts of these positions.
 To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
@@
     \centering
-    \includegraphics[width=\linewidth]{images/path_join.png}
+    \includegraphics[width=\linewidth]{images/path_join.png} %% TODO: change this
     \caption{
-        Path join of paths from the running example and the corresponding starts and ranks. The pangenome and segments are shown above to clarify the meaning of the start values.
+    Path join of paths from the running example and the corresponding starts and ranks.
+The concatenated pangenome and segments are shown above to clarify the meaning of the start values.
+\label{fig:path_join}
     }
-    \label{fig:path_join}
 \end{figure}
@@
 \subsection{Iteration}
-With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array values.
+With the previous tables stored in memory, we have all the necessary ingredients to generate the suffix array value by value.
 
@@
 \begin{itemize}
-    \item the sentinel
-    \item a separator
+    \item the sentinel $\texttt{\$}$
+    \item a separator $\texttt{\#}$
     \item a position within the last $k$ characters of a segment
@@
 
-Since the original pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.
+Since the pangenome has no corresponding position for the sentinel or separator characters, we can skip the first rows representing them.
 
@@
 The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
-Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix $P[9..] = \texttt{ACACT}$.
+Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix of a concatenated pangenome $P[9..] = \texttt{ACACT}$.
 

### 2023-08-09 manuscript/content/table_segment.tex
integrated reviewers' suggestions

\begin{table}
    \caption{
    Segment table containing an id, length, starting positions in the concatenated pangenome and ranks of the right contexts of these positions for each segment. \label{tab:segment}}
    \begin{tabular}{rrll}
        \toprule
        id & length & starts & right context ranks \\
        \midrule
        0 & 4 & 9           & 9         \\
        1 & 3 & 15, 1       & 13, 14    \\
        2 & 5 & 18, 5, 11   & 1, 2, 3   \\
        3 & 3 & 8, 14, 0    & 4, 5, 6   \\
        4 & 4 & 16          & 7         \\
        5 & 5 & 2           & 8         \\
        \bottomrule
    \end{tabular}
\end{table}

### 2023-08-09 manuscript/content/table_suffix.tex
integrated reviewers' suggestions

\begin{table*}
    \caption{
    Suffix table with the suffix array \texttt{SA}, longest common prefix array \texttt{LCP}, segment ID array \texttt{ID} and segment positions array \texttt{pos} of the running example. Suffixes of segment join from position \texttt{SA}[i] added for easier interpretation. Darker colours represent the suffix of the current segment. \label{tab:suffix}
    }
    \begin{tabular}{rrrrrl}
        \toprule
        i & \texttt{SA}[i] & \texttt{LCP}[i] & \texttt{ID}[i] & \texttt{pos}[i] & segment join[\texttt{SA}[i]..] \\
        \midrule
         0  & 30 & -1 & 6 & 0 & \texttt{\textcolor{gray}{\$}}                                     \\
         1  & 29 &  0 & 5 & 5 & \texttt{\textcolor{gray}{\#\$}}                                   \\
         2  &  4 &  1 & 0 & 4 & \texttt{\textcolor{gray}{\#ACG\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}     \\
         3  &  8 &  3 & 1 & 3 & \texttt{\textcolor{gray}{\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}          \\
         4  & 14 &  1 & 2 & 5 & \texttt{\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}                 \\
         5  & 18 &  2 & 3 & 3 & \texttt{\textcolor{gray}{\#CGAC\#CGTAC\#\$}}                      \\
         6  & 23 &  3 & 4 & 4 & \texttt{\textcolor{gray}{\#CGTAC\#\$}}                            \\
         7  & 13 &  0 & 2 & 4 & \texttt{.\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}                \\
         8  & 12 &  1 & 2 & 3 & \texttt{..\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}               \\
         9  & 27 &  0 & 5 & 3 & \texttt{AC\textcolor{gray}{\#\$}}                                 \\
        10  &  2 &  3 & 0 & 2 & \texttt{AC\textcolor{gray}{\#ACG\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}   \\
        11  & 16 &  3 & 3 & 1 & \texttt{AC\textcolor{gray}{\#CGAC\#CGTAC\#\$}}                    \\
        12  & 21 &  5 & 4 & 2 & \texttt{AC\textcolor{gray}{\#CGTAC\#\$}}                          \\
        13  &  0 &  2 & 0 & 0 & \texttt{ACAC\textcolor{gray}{\#ACG\#ACT..\#CAC\#CGAC\#CGTAC\#\$}} \\
        14  &  5 &  2 & 1 & 0 & \texttt{ACG\textcolor{gray}{\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}       \\
        15  &  9 &  2 & 2 & 0 & \texttt{ACT..\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}            \\
        16  & 28 &  0 & 5 & 4 & \texttt{C\textcolor{gray}{\#\$}}                                  \\
        17  &  3 &  2 & 0 & 3 & \texttt{C\textcolor{gray}{\#ACG\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}    \\
        18  & 17 &  2 & 3 & 2 & \texttt{C\textcolor{gray}{\#CGAC\#CGTAC\#\$}}                     \\
        19  & 22 &  4 & 4 & 3 & \texttt{C\textcolor{gray}{\#CGTAC\#\$}}                           \\
        20  &  1 &  1 & 0 & 1 & \texttt{CAC\textcolor{gray}{\#ACG\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}  \\
        21  & 15 &  4 & 3 & 0 & \texttt{CAC\textcolor{gray}{\#CGAC\#CGTAC\#\$}}                   \\
        22  &  6 &  1 & 1 & 1 & \texttt{CG\textcolor{gray}{\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}        \\
        23  & 19 &  2 & 4 & 0 & \texttt{CGAC\textcolor{gray}{\#CGTAC\#\$}}                        \\
        24  & 24 &  2 & 5 & 0 & \texttt{CGTAC\textcolor{gray}{\#\$}}                              \\
        25  & 10 &  1 & 2 & 1 & \texttt{CT..\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}             \\
        26  &  7 &  0 & 1 & 2 & \texttt{G\textcolor{gray}{\#ACT..\#CAC\#CGAC\#CGTAC\#\$}}         \\
        27  & 20 &  1 & 4 & 1 & \texttt{GAC\textcolor{gray}{\#CGTAC\#\$}}                         \\
        28  & 25 &  1 & 5 & 1 & \texttt{GTAC\textcolor{gray}{\#\$}}                               \\
        29  & 11 &  0 & 2 & 2 & \texttt{T..\textcolor{gray}{\#CAC\#CGAC\#CGTAC\#\$}}              \\
        30  & 26 &  1 & 5 & 2 & \texttt{TAC\textcolor{gray}{\#\$}}                                \\
        \bottomrule
    \end{tabular}
\end{table*}

### 2023-08-10 manuscript/content/05_conclusion.tex
added new conclusion

\section{Conclusion}
An increasing abundance of available genomic data leads to the need for new data structures to utilize all the contained information to its full potential.
Pangenome graphs have been proposed as a solution capable of representing these datasets, exploiting the repetitiveness of a collection of related genomes for efficient storage while preserving and sometimes even highlighting the underlying variations.
However, this shift in the bioinformatics field from linear sequences to pangenomic graphs brings challenges.

The enormous sizes of pangenomes make the construction of pangenomic graphs nontrivial and often one of the major bottlenecks in the analysis.
Current tools rely on well-known computational steps, such as all-against-all alignment \cite{garrison2023building}, multiple sequence alignment \cite{hickey2023pangenome}, or variant calling \cite{garrison2018variation}.
All these steps are computationally expensive and do not scale well with rapidly growing datasets.
Furthermore, variant calling relies on a linear reference, which can introduce a reference bias.

Here, we introduced the prefix-free graph as a standalone data structure and showed how to build it from currently available pangenomic datasets.
In comparison to the other tools, it offers several crucial advantages.
The construction of prefix-free graphs does not require any alignment or variant calling.
Instead, it avoids these expensive steps by employing a set of trigger words, which split the sequences according to their contexts.
The time complexity of this construction is linear with respect to the input size, and the space complexity is sublinear, proportional only to the size of the resulting data structure.

The novelty of prefix-free graphs brings several possible directions for future research, with the two main directions being the connection of prefix-free graphs to stringology and the choice of trigger words.
In Chapter 3, we showed how to use prefix-free graphs to build a suffix array of a pangenome in sublinear space.
Suffix arrays are at the core of many efficient and well-established string algorithms in bioinformatics, such as read mapping and pattern matching.
We expect this connection will facilitate the development of similar algorithms on pangenomic datasets, supporting the paradigm shift.
This direction will require exploring which additional string algorithms can use prefix-free graphs to improve their applicability to these vast and repetitive datasets.
An exciting attempt may be to map reads to prefix-free graphs in a similar fashion as the popular tool BWA \cite{li2009fast} maps reads to a linear reference.

The second direction, the choice of trigger words, is, so far, mostly undiscovered territory.
Trigger words offer great flexibility in the construction process, so a better understanding of their selection is crucial for prefix-free graphs.
From the computational perspective, a set of trigger words minimizing the size of the resulting graph is desirable since it would allow the analysis of larger datasets.
On the other hand, the construction flexibility could be used to create prefix-free graphs based on biologically significant strings.
In our experiments, we used stop codon sequences as the trigger words.
However, experiments with other motives, such as different binding sites, recombination hotspots, or repetitive elements, could illuminate the possibility of capturing particular biological phenomena with prefix-free graphs.
Moreover, integrating the strand information in the graph construction may be beneficial for further reducing the graph size and capturing biological phenomena such as inversion.
This integration could be achieved by considering both the forward and reverse complement of the trigger words.

We believe the characteristics of prefix-free graphs are highly valueable and will have a significant impact when dealing with massive datasets, moving us closer to the ultimate goal of computational pangenomics.

### 2023-08-10 manuscript/content/03_sa_construction.tex
corrected table captions

@@
 A suffix array is a permutation of string positions which lexicographically sorts the suffixes of the string starting at that position.
-It is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.
+For a set of strings, we consider a concatenation of the strings as shown in Figure \ref{fig:path_join}.
+In the following text, we will refer to this concatenation as the pangenome.
 
+The suffix array is an influential data structure with many applications in efficient string algorithms solving problems such as exact pattern matching, repeat finding, maximum exact match (MEM) finding, document retrieval and many more.
 There exist several algorithms for suffix array construction in linear time \cite{skew2003karkkainen,sais2009nong,saoverview2007puglisi} with several practical implementations \cite{louza2017inducing,libdivsufsort}.
@@
 Although they do not offer any improvement of theoretical guarantees in the worst case, in practice, they often represent the pangenome in a substantially smaller space and allow us to generate the suffix array values one by one, possibly using the values directly in subsequential computation or storing them in compressed form.
-This iteration can be done without ever expanding the pangenome to its textual representation in space proportional to the sum of segment lengths and the sum of path lengths.
+This iteration can be done without ever expanding the pangenome to its full textual representation in space proportional to the sum of segment lengths and the sum of path lengths.
 
@@
 Then, applying the ISA to these arrays changes the order of computed values in correspondence to the sorted suffixes.
-The resulting suffix array $\texttt{SA}$, longest common prefix array $\texttt{LCP}$, segment ID array $\texttt{ID}$ and segment positions array $\texttt{pos}$ are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.
+The resulting suffix array $\texttt{(SA)}$, longest common prefix array $\texttt{(LCP)}$, segment ID array $\texttt{(ID)}$ and segment positions array $\texttt{(pos)}$ are stored in a \emph{suffix table} as shown in Table \ref{tab:suffix}.
 
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth]{images/path_join.png}
+    \caption{
+    Path join of paths from the running example and the corresponding starts and ranks.
+The concatenated pangenome and segments are shown above to clarify the meaning of the start values.
+\label{fig:path_join}
+    }
+\end{figure}
+
 In the suffix table, one row can represent multiple positions of the pangenome.
-% These positions can be represented as a single integer coresponding to index in the concatenated representation of a pangenome.
 To identify these positions, we store some additional information in a \emph{segment table}.
-For each segment of the prefix-free graph, we store its length, starting positions in the concatenated pangenome and ranks of the right contexts of these positions.
+For each segment of the prefix-free graph, we store its length, starting positions in the pangenome and ranks of the right contexts of these positions.
 To calculate the starting positions and the ranks of the right contexts, we use a \emph{path join}.
@@
 An example of a path join for our running example is in Figure \ref{fig:path_join}.
-Then, starting positions can be calculated by cumulatively summing the lengths
-of the segments in path join and subtracting the overlaps.
+Then, starting positions can be calculated by cumulatively summing the lengths of the segments in path join and subtracting the overlaps.
 
@@
 Finally, we store the starts and ranks sorted by the rank values in the segment table as shown in Table \ref{tab:segment}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/path_join.png} %% TODO: change this
-    \caption{
-    Path join of paths from the running example and the corresponding starts and ranks.
-The concatenated pangenome and segments are shown above to clarify the meaning of the start values.
-\label{fig:path_join}
-    }
-\end{figure}
 
@@
 The offset from the start of a segment $\texttt{pos}[13]$ is $0$.
-Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix of a concatenated pangenome $P[9..] = \texttt{ACACT}$.
+Summing these two values, we get the first value of a suffix array $9 + 0 = 9$ corresponding to the lexicographically smallest suffix of a pangenome $P[9..] = \texttt{ACACT}$.
 
@@
 In the last case, we have a non-singleton block representing suffixes of several segments, possibly with multiple occurrences.
-These suffixes represent identical substrings in the original pangenome.
+These suffixes represent identical substrings in the pangenome.
 Here, we report a suffix array value for each of the substrings.

### 2023-08-10 manuscript/content/table_segment.tex
corrected table captions

@@
     \caption{
-    Segment table containing an id, length, starting positions in the concatenated pangenome and ranks of the right contexts of these positions for each segment. \label{tab:segment}}
+    Segment table containing an ID, length, starting positions in the pangenome and ranks of the right contexts of these positions for each segment. \label{tab:segment}}
     \begin{tabular}{rrll}

### 2023-08-10 manuscript/content/table_suffix.tex
corrected table captions

@@
     \caption{
-    Suffix table with the suffix array \texttt{SA}, longest common prefix array \texttt{LCP}, segment ID array \texttt{ID} and segment positions array \texttt{pos} of the running example.
+        Suffix table with the suffix array \texttt{(SA)}, longest common prefix array \texttt{(LCP)}, segment ID array \texttt{(ID)} and segment positions array \texttt{(pos)} of the running example.
 Suffixes of segment join from position \texttt{SA}[i] added for easier interpretation.

### 2023-08-10 manuscript/content/03_sa_construction.tex
last read done, submiting...

@@
         Segment join of segments from the running example, corresponding segment IDs, and positions.
-Permuting IDs and positions according to the \texttt{ISA} results in the segment ID array \texttt{ID} and the segment position array \texttt{pos} shown in Table \ref{tab:suffix}.
+Permuting IDs and positions according to the \texttt{ISA} results in the segment ID array \texttt{(ID)} and the segment position array \texttt{(pos)} shown in Table \ref{tab:suffix}.
 \label{fig:ids_and_positions}
